---
title: Computing GW barycenters
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Computing GW barycenters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 5, fig.height = 4
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

A GW barycenter summarizes several weighted structures with a new,
fixed-size structure. An FGW barycenter can also summarize node features. The
inputs may contain different numbers of nodes because the solver learns a
coupling from each sample to the shared barycenter.

This vignette combines three related point-cloud shapes into an eight-node
template. You will obtain its structure matrix, feature matrix, weights, and
one coupling per input sample.

```{r load}
library(rfugw)
```

## What does each sample contain?

Each sample needs a square structure matrix `C` and a feature matrix `F` with
one row per node. Here, `C` contains normalized pairwise distances, while `F`
contains two measured attributes. All feature matrices must have the same
number of columns.

```{r make-samples}
make_sample <- function(n, phase, seed) {
  set.seed(seed)
  angle <- seq(0, 2 * pi, length.out = n + 1)[-(n + 1)] + phase
  xy <- cbind(cos(angle), 0.7 * sin(angle))
  xy <- xy + matrix(rnorm(2 * n, sd = 0.04), ncol = 2)
  C <- as.matrix(dist(xy))
  C <- C / max(C)
  F <- cbind(horizontal = xy[, 1], vertical = xy[, 2])
  list(C = C, F = F)
}

samples <- list(
  make_sample(8, 0.00, 1),
  make_sample(10, 0.12, 2),
  make_sample(9, -0.10, 3)
)
Cs <- lapply(samples, `[[`, "C")
Ys <- lapply(samples, `[[`, "F")
```

```{r inspect-inputs}
data.frame(
  sample = seq_along(samples),
  nodes = vapply(Cs, nrow, integer(1)),
  features = vapply(Ys, ncol, integer(1))
)
```

Structure and feature scales affect the optimization. In this example the
structures are scaled to `[0, 1]`; the default feature-cost normalization also
scales each cross-feature cost before solving.

## How do you fit an FGW barycenter?

`fgw_barycenters()` alternates between FGW couplings and closed-form updates
of the barycenter. `N` sets the output size, not the size of any particular
sample.

```{r fit-barycenter}
bary <- fgw_barycenters(
  N = 8, Ys = Ys, Cs = Cs,
  alpha = 0.5, epsilon = 0.08,
  max_iter = 30L, tol = 1e-4, random_state = 42
)
```

```{r summarize-barycenter}
data.frame(
  status = bary$status,
  converged = bary$converged,
  nodes = nrow(bary$C),
  features = ncol(bary$X),
  iterations = bary$iterations,
  final_update = bary$error,
  objective = bary$objective
)
```

```{r check-barycenter, include = FALSE}
last <- bary$history[nrow(bary$history), ]
stopifnot(
  identical(dim(bary$C), c(8L, 8L)),
  identical(dim(bary$X), c(8L, 2L)),
  all(is.finite(bary$C)), all(is.finite(bary$X)),
  is.finite(bary$objective), is.finite(last$err_structure),
  identical(bary$status, "converged"), isTRUE(bary$converged),
  bary$error <= 1e-4,
  isTRUE(last$all_solvers_converged),
  isTRUE(last$all_inner_converged),
  is.finite(last$max_inner_residual),
  max(abs(bary$C - t(bary$C))) < 1e-10,
  max(abs(diag(bary$C))) < 1e-10,
  stats::sd(bary$C[upper.tri(bary$C)]) > 0,
  sum(apply(bary$X, 2, stats::var)) > 0,
  abs(sum(bary$p) - 1) < 1e-10
)
for (i in seq_along(bary$couplings)) {
  P <- bary$couplings[[i]]
  stopifnot(
    all(is.finite(P)), all(P >= 0), diff(range(P)) > 0,
    max(abs(rowSums(P) - bary$p)) < 1e-5,
    max(abs(colSums(P) - rep(1 / nrow(Cs[[i]]), nrow(Cs[[i]])))) < 1e-5
  )
}
```

```{r summarize-solvers}
data.frame(
  sample = seq_along(bary$solver_diagnostics),
  status = vapply(bary$solver_diagnostics, `[[`, character(1), "status"),
  inner_status = vapply(
    bary$solver_diagnostics, `[[`, character(1), "inner_status"
  ),
  max_inner_residual = vapply(
    bary$solver_diagnostics, `[[`, numeric(1), "max_inner_residual"
  )
)
```

`bary$C` is the learned `8 x 8` structure and `bary$X` is the learned `8 x 2`
feature matrix. `error` is the final update magnitude, not a proof of global
optimality. Barycenter optimization is non-convex, so inspect the history and
consider multiple initializations when the result matters.

The heat map below shows the learned pairwise structure; the feature panel
places the same eight barycenter nodes in their learned feature coordinates.

```{r plot-barycenter, echo = FALSE, fig.width = 7, fig.height = 3.2}
old_par <- par(mfrow = c(1, 2), mar = c(3, 3, 2, 1))
image(
  seq_len(nrow(bary$C)), seq_len(ncol(bary$C)), bary$C,
  xlab = "Barycenter node", ylab = "Barycenter node",
  main = "Learned structure", col = hcl.colors(20, "YlOrRd", rev = TRUE)
)
plot(
  bary$X, asp = 1, pch = 19,
  xlab = "Feature 1", ylab = "Feature 2",
  main = "Learned node features"
)
text(bary$X, labels = seq_len(nrow(bary$X)), pos = 3, cex = 0.75)
par(old_par)
```

The off-diagonal variation in `C` and the spread of `X` show that neither
learned component has collapsed to a constant.

## What do the couplings mean?

The `i`th coupling maps barycenter nodes (rows) to nodes of sample `i`
(columns). Its marginals reproduce the barycenter and sample weights.

```{r summarize-couplings}
data.frame(
  sample = seq_along(Cs),
  rows = vapply(bary$couplings, nrow, integer(1)),
  columns = vapply(bary$couplings, ncol, integer(1)),
  transported_mass = vapply(bary$couplings, sum, numeric(1))
)
```

Large entries identify strong correspondences, but `which.max()` is only a
hard summary of a soft coupling and can discard meaningful uncertainty.

## How does `alpha` change the problem?

`alpha` weights structure relative to features: `alpha = 0` is feature-only
and `alpha = 1` is structure-only. Values in between fit both. Objective values
from different `alpha` settings are not directly comparable because they weight
different losses.

```{r compare-alpha}
mostly_features <- fgw_barycenters(
  8, Ys, Cs, alpha = 0.2, epsilon = 0.08,
  max_iter = 100L, tol = 1e-4, random_state = 42
)
mostly_structure <- fgw_barycenters(
  8, Ys, Cs, alpha = 0.8, epsilon = 0.08,
  max_iter = 100L, tol = 1e-4, random_state = 42
)
```

```{r summarize-alpha-status}
data.frame(
  alpha = c(0.2, 0.8),
  status = c(mostly_features$status, mostly_structure$status),
  final_update = c(mostly_features$error, mostly_structure$error)
)
```

```{r summarize-alpha}
data.frame(
  difference = c("structure matrices", "feature matrices"),
  relative_change = c(
    sqrt(sum((mostly_structure$C - mostly_features$C)^2)) /
      sqrt(sum(mostly_features$C^2)),
    sqrt(sum((mostly_structure$X - mostly_features$X)^2)) /
      sqrt(sum(mostly_features$X^2))
  )
)
```

```{r check-alpha, include = FALSE}
alpha_changes <- c(
  sum((mostly_structure$C - mostly_features$C)^2),
  sum((mostly_structure$X - mostly_features$X)^2)
)
for (candidate in list(mostly_features, mostly_structure)) {
  candidate_last <- candidate$history[nrow(candidate$history), ]
  stopifnot(
    identical(candidate$status, "converged"), isTRUE(candidate$converged),
    candidate$error <= 1e-4,
    isTRUE(candidate_last$all_solvers_converged),
    isTRUE(candidate_last$all_inner_converged),
    is.finite(candidate_last$max_inner_residual)
  )
}
stopifnot(all(is.finite(alpha_changes)), all(alpha_changes > 0))
```

Choose `alpha` from domain knowledge or a downstream validation criterion, not
from the smallest raw objective across settings.

## How do sample weights work?

`lambdas` controls how strongly each input contributes. The values are
normalized to sum to one. This is useful when samples have unequal reliability;
it is not a substitute for node weights `ps`, which act within samples.

```{r weighted-barycenter}
weighted <- fgw_barycenters(
  8, Ys, Cs, lambdas = c(0.6, 0.25, 0.15),
  alpha = 0.5, epsilon = 0.08,
  max_iter = 30L, tol = 1e-4, random_state = 42
)
```

```{r check-weighted, include = FALSE}
stopifnot(
  all(is.finite(weighted$C)), all(is.finite(weighted$X)),
  is.finite(weighted$objective),
  sum((weighted$C - bary$C)^2) > 0
)
```

## Which barycenter function should you use?

| Available information | Function | Returned result |
|---|---|---|
| Structure and node features | `fgw_barycenters()` | Full list with structure, features, couplings, and history |
| Structure only | `entropic_gromov_barycenters()` | Structure matrix by default; full list with `log = TRUE` |
| POT-style entropic FGW interface | `entropic_fused_gromov_barycenters()` | List containing features and structure; adds diagnostics with `log = TRUE` |

You can hold one part of an FGW barycenter fixed: set `fixed_structure = TRUE`
with `init_C`, or `fixed_features = TRUE` with `init_X`. Fixing a component
changes the estimand and should reflect a substantive modeling choice.

## Where should you go next?

- `vignette("multiset-alignment")` aligns each sample to a shared fixed or
  learned template and provides projection helpers.
- `vignette("solver-guide")` explains GW and FGW formulations and their
  diagnostics.
- See `?fgw_barycenters` for initialization, inner-solver, and precision
  controls.
