---
title: "Computing GW barycenters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Computing GW barycenters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

A **Gromov-Wasserstein barycenter** is the "average shape" of a collection of
metric-measure spaces. Given several structure matrices (and optionally
features), the barycenter solver finds a new structure matrix and feature matrix
that best summarize the input collection in the GW sense.

This is useful for:

- Summarizing multiple subjects into a representative template
- Initializing a shared coordinate system for downstream alignment
- Visualizing the central tendency of a set of shapes or graphs

## Setup: three small graphs

We'll create three small "graphs" (point clouds with pairwise distance
structure) that share a similar shape but differ in detail.

```{r create-samples}
library(rfugw)

make_sample <- function(n, d_feat = 2, seed = 1) {
  set.seed(seed)
  X <- matrix(rnorm(n * 3), n, 3)
  C <- as.matrix(dist(X))
  C <- C / max(C)
  F <- matrix(rnorm(n * d_feat), n, d_feat)
  list(C = C, F = F)
}

s1 <- make_sample(12, seed = 1)
s2 <- make_sample(14, seed = 2)
s3 <- make_sample(10, seed = 3)
```

## Computing the FGW barycenter

`fgw_barycenters()` learns a fixed-size barycenter by alternating between FGW
coupling steps and closed-form barycenter updates. You specify `N`, the number
of nodes in the output barycenter.

```{r fgw-barycenter}
bary <- fgw_barycenters(
  N = 10,
  Ys = list(s1$F, s2$F, s3$F),
  Cs = list(s1$C, s2$C, s3$C),
  alpha = 0.5,
  epsilon = 0.05,
  max_iter = 50,
  random_state = 42
)
```

The result contains the learned barycenter structure and features:

```{r bary-inspect}
# Barycenter structure matrix
dim(bary$C)

# Barycenter features
dim(bary$X)

# Convergence history
tail(bary$history, 3)
```

## How alpha affects the barycenter

The `alpha` parameter controls whether the barycenter emphasizes structure
matching or feature matching:

```{r alpha-comparison}
bary_struct <- fgw_barycenters(
  N = 10,
  Ys = list(s1$F, s2$F, s3$F),
  Cs = list(s1$C, s2$C, s3$C),
  alpha = 0.9,
  epsilon = 0.05,
  max_iter = 50,
  random_state = 42
)

bary_feat <- fgw_barycenters(
  N = 10,
  Ys = list(s1$F, s2$F, s3$F),
  Cs = list(s1$C, s2$C, s3$C),
  alpha = 0.1,
  epsilon = 0.05,
  max_iter = 50,
  random_state = 42
)

cat("Structure-heavy objective:", bary_struct$objective, "\n")
cat("Feature-heavy objective:  ", bary_feat$objective, "\n")
```

## Weighted barycenters

The `lambdas` parameter lets you weight each sample differently. This is useful
when some samples are more reliable or representative:

```{r weighted-bary}
bary_w <- fgw_barycenters(
  N = 10,
  Ys = list(s1$F, s2$F, s3$F),
  Cs = list(s1$C, s2$C, s3$C),
  lambdas = c(0.5, 0.3, 0.2),
  alpha = 0.5,
  epsilon = 0.05,
  max_iter = 50,
  random_state = 42
)
bary_w$objective
```

## Fixed structure or features

You can fix the barycenter structure and only learn features (or vice versa)
by using `fixed_structure` or `fixed_features` with an initial value:

```{r fixed-structure}
# Fix the structure to sample 1's geometry, learn only features
bary_fs <- fgw_barycenters(
  N = nrow(s1$C),
  Ys = list(s1$F, s2$F, s3$F),
  Cs = list(s1$C, s2$C, s3$C),
  alpha = 0.5,
  epsilon = 0.05,
  fixed_structure = TRUE,
  init_C = s1$C,
  max_iter = 30,
  random_state = 42
)
dim(bary_fs$X)
```

## Pure GW barycenters (no features)

When you don't have features, use `gromov_barycenters()` or
`entropic_gromov_barycenters()`. These are equivalent to calling
`fgw_barycenters()` with `alpha = 1`:

```{r gw-barycenters}
# With log = FALSE (default), returns just the C matrix
bary_gw <- entropic_gromov_barycenters(
  N = 10,
  Cs = list(s1$C, s2$C, s3$C),
  epsilon = 0.05,
  max_iter = 50,
  random_state = 42
)
dim(bary_gw)
```

Set `log = TRUE` to get the full result with couplings and convergence history:

```{r gw-barycenters-log}
bary_gw_full <- entropic_gromov_barycenters(
  N = 10,
  Cs = list(s1$C, s2$C, s3$C),
  epsilon = 0.05,
  max_iter = 50,
  random_state = 42,
  log = TRUE
)
names(bary_gw_full)
```

## Examining the couplings

The barycenter result includes the final couplings from the barycenter to each
input sample. These tell you how each barycenter node corresponds to each
sample's nodes:

```{r couplings}
# Coupling from barycenter to sample 1
dim(bary$couplings[[1]])

# Best match for each barycenter node in sample 1
apply(bary$couplings[[1]], 1, which.max)
```

## Tips for larger problems

- **Warm starts**: `warmstartT = TRUE` (default) reuses previous couplings
  across outer iterations, which helps convergence.
- **Precision**: `precision = "mixed"` speeds up the inner Sinkhorn iterations.
- **Epsilon**: Larger `epsilon` converges faster but gives blurrier couplings.
  Start with 0.05--0.1 and decrease if needed.
- **N**: The barycenter size `N` doesn't need to match any input size. Choose
  it based on the resolution you need.

## Next steps

- `vignette("multiset-alignment")`: using `multialign_fit()` for multi-subject
  alignment, which offers a more scalable template-based workflow
