---
title: Linear optimal transport in rfugw
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Linear optimal transport in rfugw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 5, fig.height = 4
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Linear optimal transport matches two weighted collections when the cost of
every source-to-target match is known. You supply a cost matrix and, optionally,
weights for the two collections. `rfugw` returns a coupling: a nonnegative
matrix whose entries say how much mass moves between each pair.

This vignette builds one coupling, checks that it is trustworthy, and uses it
to map target coordinates back to the source. It then shows when to choose an
exact, entropic, or unbalanced solver.

```{r load}
library(rfugw)
```

## What are the inputs?

We will align two small point clouds in the same two-dimensional space. Rows
are points; `M[i, j]` is the Euclidean cost of matching source point `i` to
target point `j`. The weights `p` and `q` each sum to one.

```{r make-clouds}
set.seed(16)
X <- matrix(rnorm(16), ncol = 2)
Y <- matrix(rnorm(20), ncol = 2) + 0.4
M <- as.matrix(dist(rbind(X, Y)))[1:8, 9:18]
M <- M / max(M)
p <- rep(1 / nrow(X), nrow(X))
q <- rep(1 / nrow(Y), nrow(Y))
```

Linear OT requires a meaningful cross-domain cost. If your two supports live
in different metric spaces and such a cost is unavailable, use GW or FGW
instead; those methods compare within-domain structure.

## How do you compute a balanced plan?

`ot_sinkhorn()` is the usual starting point for a fast, dense coupling. Its
`epsilon` argument controls entropic regularization: larger values spread mass
more diffusely. The reported value is the unregularized transport cost
`sum(M * plan)`; it does not include the entropy penalty.

```{r fit-sinkhorn}
sink <- ot_sinkhorn(M, p, q, epsilon = 0.08, method = "auto")
data.frame(
  status = sink$status,
  cost = rfugw_value(sink),
  marginal_residual = sink$residual
)
```

```{r check-sinkhorn, include = FALSE}
P <- rfugw_plan(sink)
stopifnot(
  all(is.finite(P)), all(P >= 0),
  is.finite(rfugw_value(sink)), diff(range(P)) > 0,
  sink$residual < 1e-6
)
ot_validate_plan(sink, p, q, tol = 1e-6)
```

The marginal residual measures how closely the row and column sums reproduce
`p` and `q`. `ot_validate_plan()` provides an explicit feasibility check;
`print(sink)` and `summary(sink)` offer compact diagnostics without printing
the full coupling.

When solving a sequence of nearby regularization values, reuse the returned
dual state instead of restarting from scratch. The state has a stable gauge
(the source-weighted potential mean is zero), works across scaling and log
backends, and contains no native pointer.

```{r sinkhorn-continuation}
sink_finer <- ot_sinkhorn(
  M, p, q,
  epsilon = 0.05,
  method = "auto",
  init_duals = sink$dual_state
)
data.frame(
  status = sink_finer$status,
  initialization = sink_finer$initialization,
  iterations = sink_finer$iterations,
  residual = sink_finer$residual
)
```

You may instead pass a previous entropic plan through `init_plan`. If both are
provided, both are validated and `init_duals` takes precedence. A warm start
never carries over convergence: the new solve must satisfy its requested
marginal certificate. Because `dual_state` is an ordinary named list, it can be
saved with `saveRDS()` and reused in an installed package from a fresh session.

## How do you use the coupling?

A barycentric projection replaces each source point by the weighted average of
the target points to which it sends mass.

```{r project-points}
projected <- ot_barycentric_project(sink, Y)
head(round(projected, 3), 4)
```

```{r check-projection, include = FALSE}
stopifnot(
  identical(dim(projected), dim(X)),
  all(is.finite(projected)),
  sum(apply(projected, 2, stats::var)) > 0
)
```

```{r projection-distance}
mean_nearest_distance <- function(A, B) {
  D <- as.matrix(dist(rbind(A, B)))[
    seq_len(nrow(A)), nrow(A) + seq_len(nrow(B)), drop = FALSE
  ]
  mean(apply(D, 1, min))
}
projection_distance <- c(
  before = mean_nearest_distance(X, Y),
  after = mean_nearest_distance(projected, Y)
)
```

```{r summarize-projection}
data.frame(
  coordinates = names(projection_distance),
  mean_distance_to_nearest_target = unname(projection_distance)
)
```

```{r check-projection-distance, include = FALSE}
stopifnot(
  all(is.finite(projection_distance)),
  projection_distance[["after"]] < projection_distance[["before"]]
)
```

The result has one row per source point and the same coordinate columns as
`Y`. A source row with no transported mass cannot have a weighted average and
is returned as `NaN`; use `zero_mass = "zero"` if zeros are preferable. In
this example, the projected coordinates are closer to the target support on
average. That is an interpreted example result, not a general guarantee for
every cost or geometry.

## Which solver should you choose?

| Task | Function | Main trade-off |
|---|---|---|
| Balanced, exact linear OT | `ot_emd()` | Sparse exact plan, best suited to small or moderate problems |
| Balanced, entropic linear OT | `ot_sinkhorn()` | Faster dense plan; `epsilon` changes the solution |
| Exact transport of a specified fraction | `ot_partial_emd()` | Enforces a transported mass in `[0, 1]` |
| Entropic transport of a specified fraction | `ot_partial_sinkhorn()` | Dense fixed-mass subcoupling with certified scaling/log Dykstra |
| Exact transport with optimized mass | `ot_partial_penalized()` | A linear discard penalty selects how much mass to transport |
| Marginals may change softly | `ot_sinkhorn_unbalanced()` | KL penalties control the cost of creating or removing mass |
| Marginals change on sparse support | `ot_sinkhorn_unbalanced_ti()` | Certified TI potentials and operator application avoid a dense plan |

For example, the exact solver satisfies the same marginals without entropic
regularization.

```{r fit-exact}
exact <- ot_emd(M, p, q)
data.frame(
  converged = exact$converged,
  cost = rfugw_value(exact),
  nonzero_entries = sum(rfugw_plan(exact) > 1e-12)
)
```

```{r check-exact, include = FALSE}
stopifnot(isTRUE(exact$converged), is.finite(rfugw_value(exact)))
ot_validate_plan(exact, p, q)
```

The exact and entropic costs are useful summaries of their respective plans,
but comparing them alone is not a convergence test. Inspect the solver status
and validate the returned marginals.

## How do sparse plans remain operators?

Exact transport plans often have few nonzero edges. Converting those edges to
an `rfugw_transport_plan` preserves mass, linear application, objectives, and
barycentric projection without allocating a dense source-by-target matrix.

```{r sparse-plan}
edge_index <- which(exact$plan > 1e-12, arr.ind = TRUE)
exact_edges <- as_transport_plan(
  data.frame(
    source = edge_index[, 1],
    target = edge_index[, 2],
    weight = exact$plan[edge_index]
  ),
  n_source = nrow(exact$plan),
  n_target = ncol(exact$plan)
)
data.frame(
  representation = transport_plan_representation(exact_edges)$representation,
  stored_edges = nrow(exact_edges$data),
  transported_mass = transport_plan_mass(exact_edges),
  linear_cost = ot_linear_cost(M, exact_edges)
)
```

```{r check-sparse-plan, include = FALSE}
stopifnot(
  identical(transport_plan_shape(exact_edges),
            c(source = nrow(M), target = ncol(M))),
  abs(transport_plan_mass(exact_edges) - sum(exact$plan)) < 1e-12,
  max(abs(transport_plan_apply(exact_edges, Y) - exact$plan %*% Y)) < 1e-12,
  max(abs(transport_plan_barycentric(exact_edges, Y) -
          ot_barycentric_project(exact, Y))) < 1e-12
)
```

`transport_plan_apply()` computes the forward linear action;
`transport_plan_adjoint()` computes the transpose action. The adjoint is not an
inverse. Likewise, target-to-source barycentric projection normalizes the
adjoint by transported column mass and should not be called an inverse map.
Rows or columns with no transported mass return `NaN` by default, an explicit
abstention rather than a fabricated uniform match. Materialization is available
only by calling `transport_plan_materialize()` (or
`rfugw_plan(x, materialize = TRUE)`) explicitly. Pruning reports lost mass and
removes convergence certificates from a result until the changed plan is
separately revalidated.

## What if only a fixed fraction should match?

When the overlap is known but not all mass should participate,
`ot_partial_sinkhorn()` transports exactly the requested `mass` while treating
the supplied weights as row and column capacities. It uses the explicit
regularized objective

```text
sum(M * plan) + epsilon * sum(plan * (log(plan) - 1)).
```

This counting-measure entropy convention differs from the product-reference KL
reported by balanced `ot_sinkhorn()`, so their regularized values should not be
mixed.

```{r fit-entropic-partial}
partial_sink <- ot_partial_sinkhorn(
  M, p, q, mass = 0.7, epsilon = 0.08,
  method = "auto", tol = 1e-9
)
data.frame(
  status = partial_sink$status,
  backend = partial_sink$effective_sinkhorn_method,
  transported_mass = partial_sink$transported_mass,
  transport_term = partial_sink$transport_objective,
  entropy_term = partial_sink$weighted_entropy_minus_one,
  regularized_objective = rfugw_value(partial_sink),
  primal_dual_gap = partial_sink$duality_gap
)
```

```{r check-entropic-partial, include = FALSE}
stopifnot(
  isTRUE(partial_sink$converged), isTRUE(partial_sink$feasible),
  isTRUE(partial_sink$objective_consistent),
  isTRUE(partial_sink$objective_components_consistent),
  abs(partial_sink$transported_mass - 0.7) < 1e-9,
  max(rowSums(partial_sink$plan) - p) <=
    partial_sink$feasibility_tolerance,
  max(colSums(partial_sink$plan) - q) <=
    partial_sink$feasibility_tolerance,
  abs(partial_sink$duality_gap) <= partial_sink$duality_gap_tolerance
)
```

Auto dispatch uses bounded scaling when safe and genuine log Dykstra for large
cost-to-entropy range. A raw plan is not a valid warm start because Dykstra
also needs its correction state; pass a prior certified result or its
`warm_state` only for the identical problem.

## When is the result a Wasserstein distance?

A linear transport cost becomes a Wasserstein p-cost only after the ground
metric power is declared. `ot_wasserstein_cost()` computes
`sum(d^p * plan)`; `ot_wasserstein_distance()` returns its `1 / p` root. This
distinction prevents accidentally taking a square root of W1 or treating a
squared structural loss as a ground distance.

For supports in a common coordinate system, supply the points directly. Rows
are observations and columns are features.

```{r wasserstein-raw-supports}
w2 <- ot_wasserstein_distance(
  source = X,
  target = Y,
  p = 2,
  source_weights = p,
  target_weights = q,
  metric = "euclidean",
  solver = "exact"
)
data.frame(
  p_cost = w2$wasserstein_p_cost,
  p_distance = rfugw_value(w2),
  power = w2$wasserstein_power,
  certification = w2$value_certification
)
```

If a cost matrix already exists, declare its current power. For example, if
`M2` stores squared Euclidean distances, use `cost_power = 2`. The helper can
then transform it consistently when the requested `p` differs.

```{r wasserstein-declared-cost}
M_raw <- as.matrix(dist(rbind(X, Y)))[1:8, 9:18]
M2 <- M_raw^2
w2_from_cost <- ot_wasserstein_distance(
  cost = M2,
  cost_power = 2,
  p = 2,
  source_weights = p,
  target_weights = q
)
stopifnot(abs(rfugw_value(w2) - rfugw_value(w2_from_cost)) < 1e-10)
```

A user-supplied nonnegative matrix is accepted as a declared dissimilarity,
but `metric_certified` is `FALSE`: rfugw does not infer symmetry or the triangle
inequality from its values. The helpers normalize each positive-mass weight
vector to a probability measure and retain original/effective mass provenance.

With `solver = "sinkhorn"`, the result is labeled `estimate_kind =
"entropic_plan"`. It is the unregularized p-cost of a certified entropic plan,
not the exact Wasserstein optimum and not the entropy-regularized objective.
Nonconverged solves fail by default. `allow_uncertified = TRUE` exists only for
explicit diagnostic inspection and leaves `value_certified = FALSE`.

## What is the regularized objective and Sinkhorn divergence?

`ot_sinkhorn()` keeps `ot_dist = sum(M * plan)` for compatibility. It now also
reports the full regularized primal

```text
sum(M * plan) + epsilon * KL(plan || p %o% q),
```

plus an independently evaluated dual, their gap, and the exact constant offset
from the alternative `sum(plan * (log(plan) - 1))` entropy convention. This
matters when comparing epsilon values: the linear term alone is not the
regularized optimum, and conventions that differ by constants must not be
mixed.

```{r regularized-balanced-objective}
data.frame(
  linear_plan_cost = sink$ot_dist,
  regularized_primal = sink$regularized_objective,
  regularized_dual = sink$regularized_dual_objective,
  primal_dual_gap = sink$regularized_duality_gap
)
```

Entropic OT is biased: even a measure compared with itself generally has a
nonzero regularized value. `ot_sinkhorn_divergence()` removes that self-bias by
combining one cross solve and two self solves under the same convention.

```{r sinkhorn-divergence}
sinkhorn_div <- ot_sinkhorn_divergence(
  X, Y,
  p = 2,
  source_weights = p,
  target_weights = q,
  epsilon = 0.2,
  method = "auto",
  max_iter = 5000L,
  tol = 1e-7
)
data.frame(
  divergence = rfugw_value(sinkhorn_div),
  status = sinkhorn_div$status,
  cross = sinkhorn_div$component_values[["cross"]],
  source_self = sinkhorn_div$component_values[["source_self"]],
  target_self = sinkhorn_div$component_values[["target_self"]]
)
```

The divergence result retains the three complete component results, statuses,
residuals, iteration counts, and runtimes. It is certified only when all three
solves pass and the combined value is nonnegative within the reported numerical
tolerance. A nonconverged self solve is never hidden behind a finite cross
value.

## When should transported mass be optimized?

`ot_partial_emd()` is appropriate when the transported mass is known in
advance. When it is part of the estimand, `ot_partial_penalized()` instead
balances transport cost against a linear penalty for mass left unmatched on
either side. The supplied weights remain finite measures; they are not
separately normalized.

```{r fit-penalized-partial}
penalized_low <- ot_partial_penalized(
  M, p, q, discard_penalty = 0.05
)
penalized_high <- ot_partial_penalized(
  M, p, q, discard_penalty = 0.5
)
data.frame(
  discard_penalty = c(0.05, 0.5),
  transported_mass = c(
    penalized_low$transported_mass,
    penalized_high$transported_mass
  ),
  transport_term = c(
    penalized_low$transport_term,
    penalized_high$transport_term
  ),
  discard_term = c(
    penalized_low$discard_penalty_term,
    penalized_high$discard_penalty_term
  )
)
```

```{r check-penalized-partial, include = FALSE}
stopifnot(
  isTRUE(penalized_low$converged),
  isTRUE(penalized_high$converged),
  penalized_low$transported_mass < penalized_high$transported_mass,
  penalized_low$transported_mass > 0,
  penalized_high$transported_mass <= min(sum(p), sum(q)),
  penalized_low$objective_components_consistent,
  penalized_high$objective_components_consistent
)
```

Each transported unit avoids one source and one target discard penalty, so a
larger penalty weakly favors more transport. The result reports both discarded
masses, the transport and penalty objective terms, and an augmented exact-OT
primal/dual certificate scaled back to the original mass units. This is a
linear variable-mass formulation, not the KL marginal relaxation below.

## What if the marginals should change?

Contamination, missing support, or partial overlap can make fixed marginals a
poor model. `ot_sinkhorn_unbalanced()` replaces equality constraints with KL
penalties. Smaller `rho` permits more marginal deviation; larger `rho`
approaches the balanced problem.

```{r fit-unbalanced}
unbalanced <- ot_sinkhorn_unbalanced(
  M, p, q, epsilon = 0.08, rho = 0.2, method = "auto",
  max_iter = 1000L, tol = 1e-7, normalization = "none"
)
data.frame(
  status = unbalanced$status,
  cost = rfugw_value(unbalanced),
  regularized_objective = unbalanced$regularized_objective,
  transported_mass = unbalanced$mass
)
```

```{r check-unbalanced, include = FALSE}
Pu <- rfugw_plan(unbalanced)
stopifnot(
  all(is.finite(Pu)), all(Pu >= 0),
  is.finite(rfugw_value(unbalanced)),
  is.finite(unbalanced$mass), unbalanced$mass > 0,
  identical(unbalanced$status, "converged"),
  isTRUE(unbalanced$converged), unbalanced$residual <= 1e-7,
  diff(range(Pu)) > 0
)
```

An unbalanced plan is not expected to reproduce `p` and `q` exactly. Its total
mass and relaxed marginal residuals are part of the result and should be
interpreted alongside the application-specific choice of `rho`. Here
`normalization = "none"` means the supplied finite measures, including their
absolute and unequal total masses, enter the generalized-KL penalties unchanged.
Use `"joint"` to apply one common scale while preserving the mass ratio. The
backward-compatible 0.1 default, `"separate"`, normalizes each side to one and
should be selected only when the inputs represent probability measures. The
result records original and effective masses and separates the transport term
(`ot_dist`) from the full `regularized_objective`.

When the allowed matches themselves are sparse, use the translation-invariant
path. Here each source has only two or three admissible targets; the edge list
can represent a genuine zero cost without confusing it with a missing sparse-
matrix entry.

```{r fit-sparse-ti-uot}
ti_index <- do.call(rbind, lapply(seq_len(nrow(M)), function(i) {
  target <- unique(c(i, i + 1L, if (i == nrow(M)) ncol(M)))
  data.frame(source = i, target = target, cost = M[i, target])
}))
ti_uot <- ot_sinkhorn_unbalanced_ti(
  ti_index, p, q,
  epsilon = 0.08, rho = c(0.2, 0.3),
  n_source = nrow(M), n_target = ncol(M),
  plan = "operator", max_iter = 5000L, tol = 1e-8
)
data.frame(
  status = ti_uot$status,
  representation = ti_uot$plan_representation$representation,
  support_size = ti_uot$support_size,
  regularized_objective = rfugw_value(ti_uot),
  kkt_residual = ti_uot$kkt_residual,
  primal_dual_gap = ti_uot$primal_dual_gap
)
```

```{r check-sparse-ti-uot, include = FALSE}
ti_plan <- rfugw_plan(ti_uot)
ti_projection <- transport_plan_barycentric(
  ti_plan, X, orientation = "target_to_source"
)
stopifnot(
  isTRUE(ti_uot$converged), isTRUE(ti_uot$inner_converged),
  isTRUE(ti_uot$kkt_consistent), isTRUE(ti_uot$fixed_point_consistent),
  isTRUE(ti_uot$gauge_invariant),
  identical(transport_plan_representation(ti_plan)$representation,
            "implicit_operator"),
  identical(dim(ti_projection), dim(Y)), all(is.finite(ti_projection))
)
```

The operator applies the coupling or its adjoint from the stored cost support
and certified potentials. Request `plan = "sparse"` only when a downstream
consumer needs explicit weighted edges, or `plan = "dense"` when the dense
allocation is intentional. Unlike the older unbalanced API's legacy
`rfugw_value()` convention, the TI result's value is its full regularized
objective. Its support certificate reports both active-node coverage and a
separate exact max-flow answer to the hypothetical balanced-marginal problem;
unequal masses can make the latter false without making KL-UOT infeasible.

## How does linear OT connect to FGW?

FGW combines a cross-domain feature cost `M` with structure matrices `C1` and
`C2`. `alpha = 0` uses only the linear feature term, while `alpha = 1` uses
only structure. Intermediate values blend the two.

```{r fit-fgw}
C1 <- as.matrix(dist(X)); C1 <- C1 / max(C1)
C2 <- as.matrix(dist(Y)); C2 <- C2 / max(C2)
fgw <- fgw_entropic(
  M, C1, C2, alpha = 0.5, epsilon = 0.08, max_iter = 40L
)
data.frame(
  status = fgw$status,
  converged = fgw$converged,
  inner_status = fgw$inner_status,
  objective = fgw$fgw_dist
)
```

```{r check-fgw, include = FALSE}
fgw_value <- ot_fgw_square(M, C1, C2, fgw, alpha = 0.5)
stopifnot(
  all(is.finite(rfugw_plan(fgw))), diff(range(rfugw_plan(fgw))) > 0,
  is.finite(fgw_value), abs(fgw_value - fgw$fgw_dist) < 1e-6,
  identical(fgw$status, "converged"), isTRUE(fgw$converged),
  identical(fgw$inner_status, "converged"), isTRUE(fgw$inner_converged),
  is.finite(fgw$residual), is.finite(fgw$max_inner_residual)
)
ot_validate_plan(fgw, p, q, tol = 1e-6)
```

`ot_fgw_square()` recomputes the unregularized FGW objective from a plan. Its
agreement with the reported value checks the objective calculation; solver
status and marginal feasibility remain separate diagnostics.

## Where should you go next?

- `vignette("solver-guide")` explains how to choose among GW, partial,
  semirelaxed, and FUGW formulations.
- `vignette("barycenters")` constructs a representative metric-measure space
  from several inputs.
- See `?ot_sinkhorn`, `?ot_sinkhorn_divergence`, `?ot_emd`,
  `?ot_partial_sinkhorn`, `?ot_partial_penalized`,
  `?ot_wasserstein_distance`, and
  `?ot_validate_plan` for full argument and diagnostic contracts.
