---
title: "Linear optimal transport in rfugw"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear optimal transport in rfugw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 5, fig.height = 4
)
```

Linear optimal transport matches two weighted collections when the cost of
every source-to-target match is known. You supply a cost matrix and, optionally,
weights for the two collections. `rfugw` returns a coupling: a nonnegative
matrix whose entries say how much mass moves between each pair.

This vignette builds one coupling, checks that it is trustworthy, and uses it
to map target coordinates back to the source. It then shows when to choose an
exact, entropic, or unbalanced solver.

```{r load}
library(rfugw)
```

## What are the inputs?

We will align two small point clouds in the same two-dimensional space. Rows
are points; `M[i, j]` is the Euclidean cost of matching source point `i` to
target point `j`. The weights `p` and `q` each sum to one.

```{r make-clouds}
set.seed(16)
X <- matrix(rnorm(16), ncol = 2)
Y <- matrix(rnorm(20), ncol = 2) + 0.4
M <- as.matrix(dist(rbind(X, Y)))[1:8, 9:18]
M <- M / max(M)
p <- rep(1 / nrow(X), nrow(X))
q <- rep(1 / nrow(Y), nrow(Y))
```

Linear OT requires a meaningful cross-domain cost. If your two supports live
in different metric spaces and such a cost is unavailable, use GW or FGW
instead; those methods compare within-domain structure.

## How do you compute a balanced plan?

`ot_sinkhorn()` is the usual starting point for a fast, dense coupling. Its
`epsilon` argument controls entropic regularization: larger values spread mass
more diffusely. The reported value is the unregularized transport cost
`sum(M * plan)`; it does not include the entropy penalty.

```{r fit-sinkhorn}
sink <- ot_sinkhorn(M, p, q, epsilon = 0.08, method = "auto")
data.frame(
  status = sink$status,
  cost = rfugw_value(sink),
  marginal_residual = sink$residual
)
```

```{r check-sinkhorn, include = FALSE}
P <- rfugw_plan(sink)
stopifnot(
  all(is.finite(P)), all(P >= 0),
  is.finite(rfugw_value(sink)), diff(range(P)) > 0,
  sink$residual < 1e-6
)
ot_validate_plan(sink, p, q, tol = 1e-6)
```

The marginal residual measures how closely the row and column sums reproduce
`p` and `q`. `ot_validate_plan()` provides an explicit feasibility check;
`print(sink)` and `summary(sink)` offer compact diagnostics without printing
the full coupling.

## How do you use the coupling?

A barycentric projection replaces each source point by the weighted average of
the target points to which it sends mass.

```{r project-points}
projected <- ot_barycentric_project(sink, Y)
head(round(projected, 3), 4)
```

```{r check-projection, include = FALSE}
stopifnot(
  identical(dim(projected), dim(X)),
  all(is.finite(projected)),
  sum(apply(projected, 2, stats::var)) > 0
)
```

```{r projection-distance}
mean_nearest_distance <- function(A, B) {
  D <- as.matrix(dist(rbind(A, B)))[
    seq_len(nrow(A)), nrow(A) + seq_len(nrow(B)), drop = FALSE
  ]
  mean(apply(D, 1, min))
}
projection_distance <- c(
  before = mean_nearest_distance(X, Y),
  after = mean_nearest_distance(projected, Y)
)
```

```{r summarize-projection}
data.frame(
  coordinates = names(projection_distance),
  mean_distance_to_nearest_target = unname(projection_distance)
)
```

```{r check-projection-distance, include = FALSE}
stopifnot(
  all(is.finite(projection_distance)),
  projection_distance[["after"]] < projection_distance[["before"]]
)
```

The result has one row per source point and the same coordinate columns as
`Y`. A source row with no transported mass cannot have a weighted average and
is returned as `NaN`; use `zero_mass = "zero"` if zeros are preferable. In
this example, the projected coordinates are closer to the target support on
average. That is an interpreted example result, not a general guarantee for
every cost or geometry.

## Which solver should you choose?

| Task | Function | Main trade-off |
|---|---|---|
| Balanced, exact linear OT | `ot_emd()` | Sparse exact plan, best suited to small or moderate problems |
| Balanced, entropic linear OT | `ot_sinkhorn()` | Faster dense plan; `epsilon` changes the solution |
| Exact transport of a specified fraction | `ot_partial_emd()` | Enforces a transported mass in `[0, 1]` |
| Marginals may change softly | `ot_sinkhorn_unbalanced()` | KL penalties control the cost of creating or removing mass |

For example, the exact solver satisfies the same marginals without entropic
regularization.

```{r fit-exact}
exact <- ot_emd(M, p, q)
data.frame(
  converged = exact$converged,
  cost = rfugw_value(exact),
  nonzero_entries = sum(rfugw_plan(exact) > 1e-12)
)
```

```{r check-exact, include = FALSE}
stopifnot(isTRUE(exact$converged), is.finite(rfugw_value(exact)))
ot_validate_plan(exact, p, q)
```

The exact and entropic costs are useful summaries of their respective plans,
but comparing them alone is not a convergence test. Inspect the solver status
and validate the returned marginals.

## What if the marginals should change?

Contamination, missing support, or partial overlap can make fixed marginals a
poor model. `ot_sinkhorn_unbalanced()` replaces equality constraints with KL
penalties. Smaller `rho` permits more marginal deviation; larger `rho`
approaches the balanced problem.

```{r fit-unbalanced}
unbalanced <- ot_sinkhorn_unbalanced(
  M, p, q, epsilon = 0.08, rho = 0.2, method = "auto",
  max_iter = 1000L, tol = 1e-7
)
data.frame(
  status = unbalanced$status,
  cost = rfugw_value(unbalanced),
  transported_mass = unbalanced$mass
)
```

```{r check-unbalanced, include = FALSE}
Pu <- rfugw_plan(unbalanced)
stopifnot(
  all(is.finite(Pu)), all(Pu >= 0),
  is.finite(rfugw_value(unbalanced)),
  is.finite(unbalanced$mass), unbalanced$mass > 0,
  identical(unbalanced$status, "converged"),
  isTRUE(unbalanced$converged), unbalanced$residual <= 1e-7,
  diff(range(Pu)) > 0
)
```

An unbalanced plan is not expected to reproduce `p` and `q` exactly. Its total
mass and relaxed marginal residuals are part of the result and should be
interpreted alongside the application-specific choice of `rho`.

## How does linear OT connect to FGW?

FGW combines a cross-domain feature cost `M` with structure matrices `C1` and
`C2`. `alpha = 0` uses only the linear feature term, while `alpha = 1` uses
only structure. Intermediate values blend the two.

```{r fit-fgw}
C1 <- as.matrix(dist(X)); C1 <- C1 / max(C1)
C2 <- as.matrix(dist(Y)); C2 <- C2 / max(C2)
fgw <- fgw_entropic(
  M, C1, C2, alpha = 0.5, epsilon = 0.08, max_iter = 40L
)
data.frame(
  status = fgw$status,
  converged = fgw$converged,
  inner_status = fgw$inner_status,
  objective = fgw$fgw_dist
)
```

```{r check-fgw, include = FALSE}
fgw_value <- ot_fgw_square(M, C1, C2, fgw, alpha = 0.5)
stopifnot(
  all(is.finite(rfugw_plan(fgw))), diff(range(rfugw_plan(fgw))) > 0,
  is.finite(fgw_value), abs(fgw_value - fgw$fgw_dist) < 1e-6,
  identical(fgw$status, "converged"), isTRUE(fgw$converged),
  identical(fgw$inner_status, "converged"), isTRUE(fgw$inner_converged),
  is.finite(fgw$residual), is.finite(fgw$max_inner_residual)
)
ot_validate_plan(fgw, p, q, tol = 1e-6)
```

`ot_fgw_square()` recomputes the unregularized FGW objective from a plan. Its
agreement with the reported value checks the objective calculation; solver
status and marginal feasibility remain separate diagnostics.

## Where should you go next?

- `vignette("solver-guide")` explains how to choose among GW, partial,
  semirelaxed, and FUGW formulations.
- `vignette("barycenters")` constructs a representative metric-measure space
  from several inputs.
- See `?ot_sinkhorn`, `?ot_emd`, and `?ot_validate_plan` for full argument and
  diagnostic contracts.
