---
title: "Linear optimal transport in rfugw"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linear optimal transport in rfugw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

You have two point clouds in the same feature space and a cost for moving
mass from one support to the other. `rfugw` returns a coupling, an
unregularized transport cost, and diagnostics you can inspect before you
project points or feed the plan into a Gromov-Wasserstein solve.

```{r load}
library(rfugw)
```

## What do you have?

This guide uses two 2-d clouds. The cost `M` is pairwise Euclidean
distance. Weights `p` and `q` are the mass on each support.

```{r data}
set.seed(16)
X <- matrix(rnorm(16), 8, 2)
Y <- matrix(rnorm(20), 10, 2) + 0.4
M <- as.matrix(dist(rbind(X, Y)))[1:8, 9:18]
M <- M / max(M)
p <- rep(1 / 8, 8)
q <- rep(1 / 10, 10)
```

## Quick win: a balanced Sinkhorn plan

When both measures have the same total mass and you want a fast, dense
coupling, use `ot_sinkhorn()`. Reported `ot_dist` is `<M, plan>` and does
not add the entropic term.

```{r sinkhorn}
sink <- ot_sinkhorn(M, p, q, epsilon = 0.08)
sink$status
sink$residual
rfugw_value(sink)
```

```{r sinkhorn-check, include = FALSE}
stopifnot(is.finite(sink$ot_dist), sink$residual < 1e-6)
ot_validate_plan(sink, p, q)
```

`print(sink)` shows status and residuals without dumping the plan.
`ot_validate_plan()` checks nonnegativity and the balanced marginals.

## Which formulation should you use?

| If your data looks like this | Use | Why |
|---|---|---|
| Same total mass, both marginals known | `ot_emd()` | Exact simplex plan; small/medium problems |
| Same total mass, you want speed or a dense plan | `ot_sinkhorn()` | Entropic Sinkhorn; `method = "log"` for tiny `epsilon` |
| Mass should be created or destroyed | `ot_sinkhorn_unbalanced()` | KL penalties `rho` relax the marginals |
| Supports live in *different* metric spaces | `fgw_entropic()` / `gromov_wasserstein()` | Linear OT needs a cross-domain cost `M` |

Unregularized GW/FGW still call an exact linear-OT subproblem at each
conditional-gradient step. Entropic GW/FGW call Sinkhorn. The primitives
in this vignette are that substrate.

## Inspect convergence, then project

```{r emd-and-project}
emd <- ot_emd(M, p, q)
emd$converged
proj <- ot_barycentric_project(emd, Y)
dim(proj)
```

```{r project-check, include = FALSE}
stopifnot(isTRUE(emd$converged), nrow(proj) == nrow(X), all(is.finite(proj)))
```

Each source point maps to a weighted average of the target support.
Rows with zero transported mass become `NaN` unless you set
`zero_mass = "zero"`.

## When the masses should not match

If one cloud is contaminated or only partly overlapping, relax the
marginals with `ot_sinkhorn_unbalanced()`. Total mass can drop below 1.

```{r unbalanced}
unb <- ot_sinkhorn_unbalanced(M, p, q, epsilon = 0.08, rho = 1)
unb$status
unb$mass
```

```{r unbalanced-check, include = FALSE}
stopifnot(is.finite(unb$ot_dist), unb$mass < 1, unb$mass > 0)
```

## How this shows up in GW

Gromov-Wasserstein compares *structure* matrices `C1` and `C2` when you
do not have a shared feature cost. Fused GW mixes that structure term
with a linear cost `M` through `alpha`. The inner plan updates are the
same linear-OT primitives:

```{r gw-uses-ot}
C1 <- as.matrix(dist(X)); C1 <- C1 / max(C1)
C2 <- as.matrix(dist(Y)); C2 <- C2 / max(C2)
fgw <- fgw_entropic(M, C1, C2, alpha = 0.5, epsilon = 0.08, max_iter = 40L)
fgw$status
ot_fgw_square(M, C1, C2, fgw, alpha = 0.5)
```

```{r gw-check, include = FALSE}
stopifnot(is.finite(fgw$fgw_dist), abs(ot_fgw_square(M, C1, C2, fgw, alpha = 0.5) - fgw$fgw_dist) < 1e-6)
```

`ot_fgw_square()` recomputes the unregularized FGW objective from the
returned plan. It should match `fgw$fgw_dist`.

## Next steps

- `vignette("solver-guide")`: choose among GW, partial, semirelaxed, and FUGW
- `vignette("barycenters")`: average several metric-measure spaces
- `?ot_sinkhorn`, `?ot_emd`, `?ot_validate_plan`
- `inst/solver-contract.md`: supported names and diagnostic meanings
