---
title: "Multiset alignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiset alignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

When you have multiple subjects---brain scans, molecular graphs, sensor
networks---and want to align them all into a shared coordinate system,
`multialign_fit()` is the tool. It aligns each subject to a common template,
producing a coupling matrix per subject that maps subject nodes to template
nodes.

## Quick example

```{r quick-example}
library(rfugw)

set.seed(1)
make_subject <- function(n, d_feat = 3, seed = 1) {
  set.seed(seed)
  X <- matrix(rnorm(n * 2), n, 2)
  C <- as.matrix(dist(X))
  C <- C / max(C)
  F <- matrix(rnorm(n * d_feat), n, d_feat)
  list(C = C, F = F, id = paste0("s", seed))
}

subjects <- list(
  make_subject(15, seed = 1),
  make_subject(18, seed = 2),
  make_subject(12, seed = 3)
)

fit <- multialign_fit(
  subjects = subjects,
  k_template = 10,
  alpha = 0.5,
  epsilon = 0.05
)

names(fit$couplings)
```

Each subject gets a coupling matrix that maps its nodes to the 10-node template.

## How subjects are specified

Each subject is a list with:

- **`C`** (required): square structure matrix (e.g., pairwise distances)
- **`F`** (optional): feature matrix (`n x d`)
- **`w`** (optional): node weights (default: uniform)
- **`id`** (optional): subject identifier string

```{r subject-structure}
# All subjects can have different numbers of nodes
vapply(subjects, function(s) nrow(s$C), integer(1))
```

## Building a template

When you don't provide a template, `multialign_fit()` builds one automatically
from pooled subject features using k-means. You control the template size with
`k_template`:

```{r template-auto}
fit <- multialign_fit(
  subjects = subjects,
  template = NULL,
  k_template = 8,
  alpha = 0.5,
  epsilon = 0.05
)
dim(fit$template$C)
dim(fit$template$F)
```

You can also build the template explicitly with `multialign_make_template()`:

```{r template-explicit}
tpl <- multialign_make_template(subjects, k = 10, seed = 42)
str(tpl[c("C", "F", "w")])
```

Or provide your own template as a list with `C`, `F`, and `w`:

```{r template-custom}
fit_custom <- multialign_fit(
  subjects = subjects,
  template = tpl,
  alpha = 0.5,
  epsilon = 0.05
)
```

## Working with couplings

### Inspecting the coupling

Each coupling is an `n_subject x n_template` matrix:

```{r coupling-inspect}
P <- fit$couplings$s1
dim(P)

# Row sums should match subject weights
summary(rowSums(P))

# Column sums should match template weights
summary(colSums(P))
```

### Projecting features into a shared space

Use `multialign_project_features()` to map template-level features (e.g., a
shared embedding) back to each subject's native resolution:

```{r project-features}
# Suppose the template carries a learned embedding
E_template <- fit$template$F

# Project it to each subject's nodes
E_s1 <- multialign_project_features(fit$couplings$s1, E_template)
E_s2 <- multialign_project_features(fit$couplings$s2, E_template)

dim(E_s1)
dim(E_s2)
```

Now `E_s1` and `E_s2` live in the same d-dimensional feature space, enabling
direct cross-subject comparison.

### Projecting a subject matrix to template space

Use `multialign_project_matrix()` to push a subject's square matrix (e.g.,
connectivity, correlation) into template coordinates:

```{r project-matrix}
# Project subject 1's structure into template space
C_proj <- multialign_project_matrix(
  fit$couplings$s1,
  subjects[[1]]$C,
  normalize = "col"
)
dim(C_proj)
```

## Choosing between FGW and FUGW backends

`multialign_fit()` supports two solver backends:

- **`"fgw_entropic"`** (default): Standard Fused GW with entropic
  regularization. Both marginals are fixed.
- **`"fugw_kl"`**: Fused Unbalanced GW with KL marginal relaxation. Allows
  partial or unbalanced mass transfer.

```{r fugw-backend}
fit_fugw <- multialign_fit(
  subjects = subjects,
  k_template = 10,
  method = "fugw_kl",
  alpha = 0.5,
  epsilon = 0.01,
  reg_marginals = c(10, 10)
)

# FUGW allows non-uniform marginals
summary(rowSums(fit_fugw$couplings$s1))
```

Use `"fugw_kl"` when subject sizes differ substantially or when you expect only
partial correspondence.

## Learned templates

By default, `multialign_fit()` uses `template_mode = "fixed"`---the template
stays constant. With `template_mode = "learned"`, the template is iteratively
updated as a barycenter of the aligned subjects:

```{r learned-template}
fit_learned <- multialign_fit(
  subjects = subjects,
  template_mode = "learned",
  k_template = 10,
  alpha = 0.5,
  epsilon = 0.05,
  template_max_iter = 4,
  template_tol = 1e-6
)

# Template was refined across outer iterations
fit_learned$outer_iterations

# Convergence history
fit_learned$template_history
```

Learned templates adapt to the data but cost more computation (one full
alignment pass per outer iteration).

## Diagnostics

`multialign_fit()` returns a diagnostics data frame with per-subject solver
statistics:

```{r diagnostics}
fit$diagnostics[, c("id", "n_nodes", "objective", "iterations")]
```

## Performance tuning

### C++ batch mode

When aligning many subjects, the C++ batch kernel processes all subjects in a
single vectorized call with OpenMP parallelism:

```{r batch-mode}
fit_fast <- multialign_fit(
  subjects = subjects,
  k_template = 10,
  alpha = 0.5,
  epsilon = 0.05,
  use_cpp_batch = TRUE,
  n_threads = 2
)
fit_fast$used_cpp_batch
```

### Mixed precision

`precision = "mixed"` (the default in `multialign_fit()`) uses single-precision
inner Sinkhorn iterations for a 1.5--2x speedup:

```{r mixed-precision}
fit_mixed <- multialign_fit(
  subjects = subjects,
  k_template = 10,
  alpha = 0.5,
  epsilon = 0.05,
  precision = "mixed"
)
```

### Autotune

For large problems, `autotune = TRUE` automatically caps iteration counts and
check intervals based on problem size:

```{r autotune}
fit_auto <- multialign_fit(
  subjects = subjects,
  k_template = 10,
  alpha = 0.5,
  epsilon = 0.05,
  autotune = TRUE,
  autotune_level = "moderate"
)
fit_auto$autotuned_controls
```

### Low-rank structure approximation

`structure_rank` enables a low-rank approximation of the GW tensor product.
Use `"auto"` for size-aware rank selection:

```{r structure-rank}
fit_lr <- multialign_fit(
  subjects = subjects,
  k_template = 10,
  alpha = 0.5,
  epsilon = 0.05,
  structure_rank = "auto"
)
fit_lr$structure_rank
```

## Putting it together: a typical workflow

```{r full-workflow}
# 1. Prepare subjects
subjects <- lapply(1:5, function(i) {
  make_subject(n = sample(10:20, 1), seed = i)
})

# 2. Align to a shared template
fit <- multialign_fit(
  subjects = subjects,
  k_template = 12,
  method = "fgw_entropic",
  alpha = 0.5,
  epsilon = 0.05,
  precision = "mixed",
  use_cpp_batch = TRUE,
  n_threads = 2
)

# 3. Project all subjects into the shared feature space
shared_features <- lapply(names(fit$couplings), function(id) {
  multialign_project_features(fit$couplings[[id]], fit$template$F)
})

# 4. Now all subjects have features in the same 3-d space
vapply(shared_features, dim, integer(2))
```

## Next steps

- `vignette("solver-guide")`: details on each solver variant
- `vignette("barycenters")`: computing GW barycenters directly
