---
title: "Multiset alignment"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiset alignment}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE,
  fig.width = 5, fig.height = 4
)
```

Multiset alignment puts several attributed structures into correspondence with
one shared template. A subject might be a brain parcellation, graph, or point
cloud; subjects may have different numbers of nodes. `multialign_fit()` returns
one subject-to-template coupling per subject, plus the template and solver
diagnostics.

This vignette aligns three related point clouds, checks their couplings, then
uses those couplings to move features and square matrices between subject and
template coordinates.

```{r load}
library(rfugw)
```

## What does a subject look like?

Each subject is a list containing a square structure matrix `C`. It may also
contain node features `F`, probability weights `w`, and a character `id`.
Features must use the same columns across subjects.

```{r make-subjects}
make_subject <- function(n, phase, seed) {
  set.seed(seed)
  angle <- seq(0, 2 * pi, length.out = n + 1)[-(n + 1)] + phase
  xy <- cbind(cos(angle), 0.65 * sin(angle))
  xy <- xy + matrix(rnorm(2 * n, sd = 0.04), ncol = 2)
  C <- as.matrix(dist(xy))
  F <- cbind(horizontal = xy[, 1], vertical = xy[, 2])
  list(C = C / max(C), F = F, id = paste0("subject_", seed))
}

subjects <- list(
  make_subject(12, 0.00, 1),
  make_subject(15, 0.12, 2),
  make_subject(10, -0.10, 3)
)
```

```{r inspect-subjects}
data.frame(
  id = vapply(subjects, `[[`, character(1), "id"),
  nodes = vapply(subjects, function(x) nrow(x$C), integer(1)),
  features = vapply(subjects, function(x) ncol(x$F), integer(1))
)
```

Missing `w` means uniform node weights. A missing `F` is allowed only when a
compatible template is supplied; automatic template construction pools the
subject features.

## How do you align to a shared template?

With `template = NULL`, `multialign_fit()` builds a template by clustering the
pooled features. `k_template` chooses its number of nodes. The default
`fgw_entropic` backend balances feature similarity and structural similarity.

```{r fit-alignment}
fit <- multialign_fit(
  subjects,
  k_template = 8,
  alpha = 0.5,
  epsilon = 0.08,
  max_iter = 50L,
  sinkhorn_max_iter = 250L,
  tol = 1e-9
)
```

```{r summarize-alignment}
data.frame(
  id = names(fit$couplings),
  subject_nodes = vapply(fit$couplings, nrow, integer(1)),
  template_nodes = vapply(fit$couplings, ncol, integer(1)),
  status = fit$diagnostics$status,
  converged = fit$diagnostics$converged,
  inner_status = fit$diagnostics$inner_status,
  objective = fit$objectives
)
```

```{r check-alignment, include = FALSE}
stopifnot(
  identical(names(fit$couplings), vapply(subjects, `[[`, character(1), "id")),
  identical(dim(fit$template$C), c(8L, 8L)),
  all(is.finite(fit$template$C)), all(is.finite(fit$template$F)),
  stats::sd(fit$template$C[upper.tri(fit$template$C)]) > 0,
  all(is.finite(fit$objectives)),
  all(is.finite(fit$diagnostics$error)),
  all(fit$diagnostics$error <= 1e-9),
  all(fit$diagnostics$status == "outer_converged_inner_unavailable"),
  all(is.na(fit$diagnostics$converged)),
  all(fit$diagnostics$inner_status == "unavailable"),
  all(is.na(fit$diagnostics$inner_converged))
)
for (i in seq_along(subjects)) {
  P <- fit$couplings[[i]]
  subject_w <- rep(1 / nrow(subjects[[i]]$C), nrow(subjects[[i]]$C))
  stopifnot(
    all(is.finite(P)), all(P >= 0), diff(range(P)) > 0,
    max(abs(rowSums(P) - subject_w)) < 2e-5,
    max(abs(colSums(P) - fit$template$w)) < 2e-5
  )
}
```

Each coupling has one row per subject node and one column per template node.
For balanced FGW, its row and column sums reproduce the subject and template
weights. Different node counts do not by themselves require an unbalanced
solver. This run uses the native batch path: the outer stopping residual and
balanced feasibility checks pass, but the batch kernel does not return nested
Sinkhorn certificates. Consequently `converged` is `NA` and `inner_status` is
`"unavailable"`; the result does not claim fully certified convergence.

## How do you reuse a template?

Build a template explicitly when several analyses should use the same target,
or when you want to inspect and save it before fitting.

```{r make-template}
template <- multialign_make_template(subjects, k = 8, seed = 42)
fit_fixed <- multialign_fit(
  subjects, template = template,
  alpha = 0.5, epsilon = 0.08,
  max_iter = 50L, sinkhorn_max_iter = 250L
)
```

```{r summarize-fixed}
fit_fixed$diagnostics[, c(
  "id", "status", "converged", "inner_status", "objective"
)]
```

```{r check-fixed, include = FALSE}
stopifnot(
  identical(fit_fixed$template, template),
  length(fit_fixed$couplings) == length(subjects),
  all(is.finite(fit_fixed$objectives)),
  all(fit_fixed$diagnostics$status ==
        "outer_converged_inner_unavailable"),
  all(is.na(fit_fixed$diagnostics$converged)),
  all(fit_fixed$diagnostics$inner_status == "unavailable"),
  all(is.na(fit_fixed$diagnostics$inner_converged)),
  all(is.finite(fit_fixed$diagnostics$error)),
  all(fit_fixed$diagnostics$error <= 1e-9)
)
for (i in seq_along(subjects)) {
  P <- fit_fixed$couplings[[i]]
  subject_w <- rep(1 / nrow(subjects[[i]]$C), nrow(subjects[[i]]$C))
  stopifnot(
    max(abs(rowSums(P) - subject_w)) < 2e-5,
    max(abs(colSums(P) - template$w)) < 2e-5
  )
}
```

An explicit template can also be any valid list with `C`, `F`, and `w`. Reusing
one makes coordinates comparable across separate runs; an automatically built
template is specific to the subjects supplied in that call.

## How do you project template features to a subject?

`multialign_project_features()` maps a template-level matrix back to subject
resolution. Row normalization turns each subject row of the coupling into
weights over template nodes.

```{r project-features}
P1 <- fit$couplings$subject_1
subject_1_features <- multialign_project_features(P1, fit$template$F)
data.frame(
  feature = colnames(subject_1_features),
  correlation = diag(cor(subjects[[1]]$F, subject_1_features))
)
```

```{r check-features, include = FALSE}
stopifnot(
  identical(dim(subject_1_features), c(12L, 2L)),
  all(is.finite(subject_1_features)),
  sum(apply(subject_1_features, 2, stats::var)) > 0,
  all(diag(cor(subjects[[1]]$F, subject_1_features)) > 0.9)
)
```

The output has one row per subject node and the same columns as the template
features. The high correlations show that this synthetic alignment preserves
the known feature pattern. Because the template was built from these same
features, this is an illustrative sanity check, not independent accuracy
evidence.

## How do you move a square matrix to template space?

`multialign_project_matrix()` pushes a subject-level square matrix through both
sides of the coupling. This is useful for connectivity or similarity matrices.

```{r project-matrix}
template_structure <- multialign_project_matrix(
  P1, subjects[[1]]$C, normalize = "col"
)
dim(template_structure)
```

```{r check-matrix, include = FALSE}
stopifnot(
  identical(dim(template_structure), c(8L, 8L)),
  all(is.finite(template_structure)),
  max(abs(template_structure - t(template_structure))) < 1e-10,
  stats::sd(template_structure[upper.tri(template_structure)]) > 0
)
```

Normalization changes the interpretation. Use `"row"` when interpolating
template values at subject nodes and `"col"` when aggregating subject values
at template nodes; use `multialign_normalize_plan()` directly when you need to
inspect those weights.

## When should you use unbalanced FUGW?

The default `method = "fgw_entropic"` fixes both marginals. Choose
`method = "fugw_kl"` when unmatched regions, contamination, or genuine mass
variation make exact marginal preservation inappropriate.

```{r fit-unbalanced}
fit_unbalanced <- multialign_fit(
  subjects,
  template = template,
  method = "fugw_kl",
  alpha = 0.5,
  epsilon = 0.5,
  reg_marginals = c(1, 1),
  max_iter = 200L,
  tol = 1e-6,
  max_iter_ot = 1000L,
  tol_ot = 1e-6,
  use_cpp_batch = FALSE
)
data.frame(
  id = names(fit_unbalanced$couplings),
  status = fit_unbalanced$diagnostics$status,
  inner_status = fit_unbalanced$diagnostics$inner_status,
  transported_mass = fit_unbalanced$diagnostics$plan_mass
)
```

```{r check-unbalanced, include = FALSE}
unbalanced_mass <- vapply(fit_unbalanced$couplings, sum, numeric(1))
stopifnot(
  all(vapply(fit_unbalanced$couplings, function(P) {
    all(is.finite(P)) && all(P >= 0) && diff(range(P)) > 0
  }, logical(1))),
  all(is.finite(unbalanced_mass)), all(unbalanced_mass > 0),
  all(fit_unbalanced$diagnostics$status == "converged"),
  all(fit_unbalanced$diagnostics$converged),
  all(fit_unbalanced$diagnostics$inner_status == "converged"),
  all(fit_unbalanced$diagnostics$inner_converged),
  all(is.finite(fit_unbalanced$diagnostics$error)),
  all(fit_unbalanced$diagnostics$error <= 1e-6),
  all(is.finite(fit_unbalanced$diagnostics$max_inner_residual)),
  all(fit_unbalanced$diagnostics$max_inner_residual <= 1e-6)
)
```

Unbalanced row and column sums are intentionally relaxed. The
`reg_marginals` penalties control that relaxation, so transported mass and
solver diagnostics should be checked rather than judged by balanced marginal
equalities.

## Should the template be fixed or learned?

The default `template_mode = "fixed"` estimates couplings to an unchanged
template. `template_mode = "learned"` alternates alignment with
barycenter-style template updates.

```{r fit-learned-template}
learned <- multialign_fit(
  subjects,
  template = template,
  template_mode = "learned",
  alpha = 0.5,
  epsilon = 0.08,
  max_iter = 40L,
  template_max_iter = 3L
)
last_template_step <- learned$template_history[learned$outer_iterations, ]
data.frame(
  outer_iterations = learned$outer_iterations,
  stopped_at_cap = learned$outer_iterations == 3L,
  template_tolerance_met =
    last_template_step$delta_template < 1e-6 &&
    last_template_step$delta_objective < 1e-6
)
```

```{r check-learned, include = FALSE}
stopifnot(
  is.data.frame(learned$template_history),
  nrow(learned$template_history) == learned$outer_iterations,
  all(is.finite(learned$template_history$objective_total)),
  all(is.finite(learned$template_history$delta_template)),
  learned$outer_iterations == 3L,
  !(last_template_step$delta_template < 1e-6 &&
      last_template_step$delta_objective < 1e-6),
  max(abs(learned$template$C - template$C)) > 0
)
```

A learned template is a fitted summary of the supplied collection rather than
a fixed external reference. It costs additional alignment passes and makes the
result specific to those inputs. In this example the run reaches its
three-iteration cap; it must not be described as converged unless both reported
deltas meet `template_tol`. Even when they do, the non-convex optimization does
not establish a unique or globally optimal template.

## What should you tune first?

- `alpha` controls structure (`1`) versus features (`0`). Choose it from the
  scientific question or downstream validation.
- `epsilon` controls entropic smoothing. Smaller values can sharpen couplings
  but are harder to optimize.
- `precision`, batching, low-rank structure, and `autotune` are performance
  controls. Establish a trustworthy full-precision result before relying on
  approximations, and compare both objectives and coupling feasibility.

## Where should you go next?

- `vignette("barycenters")` explains direct fixed-support GW and FGW
  barycenters.
- `vignette("solver-guide")` covers solver formulations, convergence, and
  diagnostics.
- See `?multialign_fit` and `?multialign_project_features` for the complete
  interfaces.
