---
title: "Getting started with rfugw"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rfugw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

You have two collections of points---maybe brain regions, molecular structures,
or sensor networks---and you want to find the best correspondence between them.
The catch: the two collections live in completely different coordinate systems,
so you can't just overlay them. All you have is the *internal geometry* of each
collection (which points are close to which) and, optionally, some features
attached to each point.
**Gromov-Wasserstein (GW)** and **Fused Gromov-Wasserstein (FGW)** solve exactly
this problem. `rfugw` gives you fast C++ implementations of both, callable from R.

## Quick example

```{r quick-example}
library(rfugw)

set.seed(42)
# Two small point clouds with different sizes
src <- matrix(rnorm(20 * 2), 20, 2)
tgt <- matrix(rnorm(25 * 2), 25, 2)

# Internal geometry: pairwise distance matrices
C1 <- as.matrix(dist(src))
C2 <- as.matrix(dist(tgt))

# Normalize to [0, 1] for numerical stability
C1 <- C1 / max(C1)
C2 <- C2 / max(C2)

# Find the optimal structural coupling
out <- entropic_gromov_wasserstein(C1, C2, epsilon = 0.05)
str(out[c("gw_dist", "iterations")])
```

The `plan` matrix tells you how mass flows from each source point to each target
point---it's the soft correspondence.

## What are these matrices?

`rfugw` works with three inputs:

- **C1** (`ns x ns`): structure matrix for the source. Usually pairwise
  distances, but any symmetric cost matrix works.
- **C2** (`nt x nt`): structure matrix for the target. Same idea, different size.
- **M** (`ns x nt`): cross-domain feature cost. Only needed for *Fused* GW.
  Measures how dissimilar source point *i*'s features are from target point *j*'s.

The solver finds a coupling (transport plan) that simultaneously respects the
internal structure of both domains and (when `M` is provided) the feature
similarity.

## Gromov-Wasserstein: structure only

When you only have geometry---no shared features---use `gromov_wasserstein()`
(exact, via conditional gradient) or `entropic_gromov_wasserstein()` (faster,
regularized).

```{r gw-solve}
# Exact GW via conditional gradient
out_exact <- gromov_wasserstein(C1, C2)
out_exact$gw_dist
```

```{r gw-entropic}
# Entropic GW (faster for larger problems)
out_ent <- entropic_gromov_wasserstein(C1, C2, epsilon = 0.05)
out_ent$gw_dist
```

The entropic version adds a small entropy penalty (`epsilon`) that smooths the
optimization landscape. Smaller `epsilon` gives sharper couplings but converges
more slowly.

## Fused Gromov-Wasserstein: structure + features

When each point also carries features (e.g., signal amplitude, color, gene
expression), FGW blends structure matching with feature matching. The `alpha`
parameter controls the trade-off: `alpha = 1` ignores features (pure GW),
`alpha = 0` ignores structure (pure Wasserstein on features).

```{r fgw-setup}
# Attach 3-dimensional features to each point
F_src <- matrix(rnorm(20 * 3), 20, 3)
F_tgt <- matrix(rnorm(25 * 3), 25, 3)

# Cross-domain feature cost (Euclidean distances between features)
M <- as.matrix(dist(rbind(F_src, F_tgt)))[1:20, 21:45]
M <- M / max(M)
```

```{r fgw-solve}
out_fgw <- fgw_entropic(
  M, C1, C2,
  alpha = 0.5,
  epsilon = 0.05
)
out_fgw$fgw_dist
```

```{r fgw-plan-inspect}
# The coupling is a 20 x 25 matrix; look at its structure
dim(out_fgw$plan)
range(out_fgw$plan)
```

## Inspecting the coupling

The transport plan has row sums equal to the source weights and column sums
equal to the target weights (uniform by default).

```{r plan-marginals}
# Row marginals (source)
summary(rowSums(out_fgw$plan))

# Column marginals (target)
summary(colSums(out_fgw$plan))
```

To find the best-matching target point for each source point, take the column
index of the maximum in each row:

```{r hard-assignment}
best_match <- apply(out_fgw$plan, 1, which.max)
head(best_match)
```

## Key parameters

| Parameter | What it does | Typical range |
|-----------|-------------|---------------|
| `epsilon` | Entropic regularization. Larger = smoother, faster; smaller = sharper. | 0.001--0.1 |
| `alpha` | Feature vs. structure weight (FGW only). 1 = pure GW. | 0--1 |
| `max_iter` | Outer iteration limit. | 100--1000 |
| `precision` | `"mixed"` (faster) or `"double"` (more accurate). | --- |
| `solver` | `"PGD"` (projected gradient) or `"PPA"` (proximal point). | --- |

## The `2` suffix convention

Every solver has a companion function with a `2` suffix (e.g.,
`fgw_entropic2()`, `gromov_wasserstein2()`) that returns only the scalar
objective value instead of the full result list. Use these when you only need
the distance:
```{r two-suffix}
fgw_entropic2(M, C1, C2, alpha = 0.5, epsilon = 0.05)
```

## Next steps

- `vignette("solver-guide")`: choosing the right solver variant
  (partial, semirelaxed, unbalanced)
- `vignette("barycenters")`: computing average shapes with GW barycenters
- `vignette("multiset-alignment")`: aligning many subjects to a shared template
