---
title: Getting started with rfugw
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Getting started with rfugw}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Suppose you have two collections of objects---brain regions from two subjects,
for example---but their coordinates are not directly comparable. You do know
the distances *within* each collection, and you may have features measured in
both. `rfugw` turns those inputs into a transport plan: a matrix of soft
correspondences between the collections.

This vignette follows one small alignment from data to result. It introduces
the default entropic fused Gromov--Wasserstein (FGW) workflow first, then shows
how to omit features and how to inspect the result safely.

## What data do you need?

Each row below is one object. The target is a reordered, rotated version of the
source, so ordinary coordinate distance between the two matrices would be
misleading.

```{r example-data}
library(rfugw)

source_xy <- rbind(A = c(0, 0), B = c(1, 0), C = c(1.4, 0.8),
                   D = c(0.7, 1.5), E = c(-0.3, 0.9), F = c(0.2, 0.4))
target_order <- c(4, 1, 6, 3, 5, 2)
```

```{r make-target}
theta <- pi / 3
rotation <- matrix(c(cos(theta), -sin(theta),
                     sin(theta),  cos(theta)), 2, 2)
target_xy <- source_xy[target_order, ] %*% rotation
target_xy <- sweep(target_xy, 2, c(4, -2), "+")
rownames(target_xy) <- rownames(source_xy)[target_order]
```

GW and FGW use three cost matrices:

- `C1` is the square structure cost within the source;
- `C2` is the square structure cost within the target;
- `M` is the rectangular feature cost from source rows to target rows.

Pairwise distances are a common choice for `C1` and `C2`. Here each object also
has a shared scalar feature; `M[i, j]` is the absolute feature difference for a
possible match.

```{r build-costs}
C1 <- as.matrix(dist(source_xy))
C2 <- as.matrix(dist(target_xy))
C1 <- C1 / max(C1); C2 <- C2 / max(C2)

source_feature <- c(0.05, 0.25, 0.50, 0.70, 0.90, 0.38)
target_feature <- source_feature[target_order]
M <- abs(outer(source_feature, target_feature, "-"))
```

Scaling comparable costs to similar ranges makes `alpha` easier to interpret.
Cost matrices must be finite; structure matrices must be square, while `M`
must have `nrow(C1)` rows and `nrow(C2)` columns.

```{r validate-inputs, include = FALSE}
stopifnot(
  all(is.finite(C1)), all(is.finite(C2)), all(is.finite(M)),
  all(dim(C1) == c(6, 6)), all(dim(C2) == c(6, 6)),
  all(dim(M) == c(6, 6)), max(C1) > 0, max(C2) > 0, max(M) > 0
)
```

## How do you fit an alignment?

Use `fgw_entropic()` when both structure and shared features should influence
the match. `alpha` is the structure share: `alpha = 1` uses structure alone,
and `alpha = 0` uses features alone. Positive `epsilon` smooths the plan; larger
values generally spread mass more broadly.

```{r fit-fgw}
fit <- fgw_entropic(
  M, C1, C2, alpha = 0.5,
  epsilon = 0.08, sinkhorn_tol = 1e-8
)
fit
```

The compact print method reports the formulation, backend, status, objective,
iteration count, and residual without printing the whole plan. Optimization is
non-convex, so this is a solver result, not a certificate of the global optimum.

```{r check-fit, include = FALSE}
plan <- rfugw_plan(fit)
p <- rep(1 / nrow(C1), nrow(C1))
q <- rep(1 / nrow(C2), nrow(C2))
ot_validate_plan(fit, p, q, tol = 1e-6)
stopifnot(
  all(is.finite(plan)), all(plan >= 0), diff(range(plan)) > 0,
  is.finite(rfugw_value(fit)),
  identical(rfugw_status(fit), "converged"),
  isTRUE(rfugw_residuals(fit)$inner_converged),
  max(abs(rowSums(plan) - p)) < 1e-6,
  max(abs(colSums(plan) - q)) < 1e-6
)
```

## What does the plan say?

Rows of the plan correspond to source objects and columns to target objects.
A balanced plan has the requested source and target weights as its row and
column sums. For a compact view, report the strongest target for each source.

```{r summarize-matches}
best_column <- apply(rfugw_plan(fit), 1, which.max)
matches <- data.frame(
  source = rownames(source_xy),
  target = rownames(target_xy)[best_column]
)
matches
```

In this constructed example, the recovered target labels agree with the source
labels. In real analyses there may be no one-to-one answer: keep the full soft
plan when its uncertainty or mass splitting matters.

```{r validate-matches, include = FALSE}
stopifnot(identical(matches$source, matches$target))
```

## What if you have structure but no shared features?

Use `entropic_gromov_wasserstein()` when only within-domain structure is
meaningful. Its input is just the two structure matrices.

```{r fit-gw}
gw_fit <- entropic_gromov_wasserstein(
  C1, C2, epsilon = 0.08, sinkhorn_tol = 1e-8
)
summary(gw_fit)
```

The result has the same central contract: `rfugw_plan()` extracts the coupling,
`rfugw_value()` extracts the documented objective, and `rfugw_status()` reports
the stopping status.

```{r check-gw, include = FALSE}
gw_plan <- rfugw_plan(gw_fit)
ot_validate_plan(gw_fit, p, q, tol = 1e-6)
stopifnot(
  all(is.finite(gw_plan)), diff(range(gw_plan)) > 0,
  is.finite(rfugw_value(gw_fit)),
  identical(rfugw_status(gw_fit), "converged"),
  isTRUE(rfugw_residuals(gw_fit)$inner_converged)
)
```

Unregularized `gromov_wasserstein()` and `fgw_exact_cg()` are alternatives when
you specifically want conditional-gradient optimization with exact linear-OT
directions. They can produce sharper plans, but the outer GW/FGW problem remains
non-convex.

## Which result fields should you trust?

Prefer the stable accessors over reaching into solver-specific lists:

- `rfugw_plan()` returns the transport plan;
- `rfugw_value()` returns the documented objective;
- `rfugw_status()` returns the solver status;
- `rfugw_residuals()` returns feasibility and stopping diagnostics.

Always inspect status and residuals before interpreting a plan. A finite plan
can still come from a run that reached an iteration limit. Also compare
objective values only when the formulation, inputs, weights, and regularization
are the same; an entropically regularized value is not interchangeable with an
unregularized GW value.

Several solver families provide a documented scalar companion, such as
`fgw_entropic2()` and `gromov_wasserstein2()`. These return only the objective
and are convenient when the coupling and diagnostics are not needed.

## Where should you go next?

- `vignette("solver-guide")` explains how marginal assumptions determine the
  solver family.
- `vignette("linear-ot")` covers transport when you already have a direct
  cross-domain cost matrix.
- `vignette("barycenters")` constructs representative metric spaces.
- `vignette("multiset-alignment")` aligns several collections to a template.
