---
title: Choosing the right GW solver
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Choosing the right GW solver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Choosing a solver starts with the correspondence you want. Do both
distributions have fixed mass? Should every object participate? Do you have
features that can be compared across domains? The answers determine the model;
regularization and numerical controls come afterward.

This guide fits several formulations to the same two small metric spaces. Each
returns a soft correspondence, but the plans obey different marginal
constraints and their objective values are not directly comparable.

Different numbers of source and target objects do not, by themselves, require
partial or unbalanced transport. Balanced transport handles rectangular plans;
choose a relaxed formulation only when the mass assumptions call for one.

## What are we aligning?

The source and target contain different numbers of objects. Each has an
internal distance matrix, and each object has two features measured in a common
space.

```{r example-data}
library(rfugw)

source_xy <- rbind(c(0, 0), c(1, 0), c(1.2, 0.8), c(0.4, 1.3), c(-0.3, 0.6))
target_xy <- rbind(c(2.0, 1.0), c(2.5, 1.8), c(1.9, 2.4),
                   c(1.2, 2.0), c(1.0, 1.2), c(2.8, 2.6))
C1 <- as.matrix(dist(source_xy)); C1 <- C1 / max(C1)
C2 <- as.matrix(dist(target_xy)); C2 <- C2 / max(C2)
```

```{r feature-cost}
F1 <- cbind(source_xy[, 1] + source_xy[, 2], source_xy[, 2])
F2 <- cbind(target_xy[, 1] + target_xy[, 2] - 3, target_xy[, 2] - 1)
M <- as.matrix(dist(rbind(F1, F2)))[1:5, 6:11]
M <- M / max(M)
p <- rep(1 / nrow(C1), nrow(C1))
q <- rep(1 / nrow(C2), nrow(C2))
```

```{r validate-data, include = FALSE}
stopifnot(
  all(is.finite(C1)), all(is.finite(C2)), all(is.finite(M)),
  all(dim(C1) == c(5, 5)), all(dim(C2) == c(6, 6)),
  all(dim(M) == c(5, 6)),
  max(C1) > 0, max(C2) > 0, max(M) > 0
)
```

## Which family matches your assumptions?

Use the narrowest formulation that expresses the scientific problem:

| Your data and assumptions | Solver family |
|---|---|
| A direct cost between objects; no relational structure | `ot_sinkhorn()` or `ot_emd()` |
| Structure only; both marginals fixed | `entropic_gromov_wasserstein()` |
| Structure plus shared features; both marginals fixed | `fgw_entropic()` |
| Only a specified amount of mass should match | `entropic_partial_*()` or `partial_*()` |
| Source weights fixed; target weights learned | `entropic_semirelaxed_*()` or `semirelaxed_*()` |
| Both marginals may depart from reference weights | `fugw_kl()` |

Within the balanced GW and FGW families, the unregularized counterparts
`gromov_wasserstein()` and `fgw_exact_cg()` use conditional gradient with exact
linear-OT directions. The entropic functions are usually the practical first
attempt because their positive regularization produces smoother plans and
avoids an exact transport solve at every outer iteration.

## When are both marginals fixed?

Balanced FGW requires every row and column sum to match the supplied weights.
Use it when both distributions represent complete probability measures and all
mass should be assigned.

```{r balanced-fit}
balanced <- fgw_entropic(
  M, C1, C2, p = p, q = q,
  alpha = 0.5, epsilon = 0.08
)
balanced
```

```{r check-balanced, include = FALSE}
balanced_plan <- rfugw_plan(balanced)
ot_validate_plan(balanced, p, q, tol = 1e-6)
stopifnot(
  all(is.finite(balanced_plan)), diff(range(balanced_plan)) > 0,
  is.finite(rfugw_value(balanced)),
  identical(rfugw_status(balanced), "converged"),
  isTRUE(balanced$converged),
  isTRUE(rfugw_residuals(balanced)$inner_converged),
  max(abs(rowSums(balanced_plan) - p)) < 1e-6,
  max(abs(colSums(balanced_plan) - q)) < 1e-6
)
```

Set `alpha = 1` or call `entropic_gromov_wasserstein(C1, C2)` when no
cross-domain features are available. Do not compare objective values across
different `alpha` or `epsilon` settings as though they were the same quantity.

## When should only part of the mass match?

Partial GW is for overlap known through a total transported mass `m`. Rows and
columns may retain unmatched mass, but the plan must transport exactly `m` and
must not exceed either marginal.

```{r partial-fit}
partial <- entropic_partial_gromov_wasserstein(
  C1, C2, p = p, q = q,
  m = 0.7, reg = 0.10, log = TRUE
)
data.frame(
  status = rfugw_status(partial),
  objective = rfugw_value(partial),
  transported_mass = sum(rfugw_plan(partial))
)
```

```{r check-partial, include = FALSE}
partial_plan <- rfugw_plan(partial)
ot_validate_plan(
  partial_plan, p, q, mass = 0.7,
  marginals = "partial", tol = 1e-6
)
stopifnot(
  all(is.finite(partial_plan)), diff(range(partial_plan)) > 0,
  is.finite(rfugw_value(partial)),
  identical(rfugw_status(partial), "converged"),
  isTRUE(rfugw_residuals(partial)$inner_converged),
  abs(sum(partial_plan) - 0.7) < 1e-6
)
```

The exact `partial_gromov_wasserstein()` and
`partial_fused_gromov_wasserstein()` default to the built-in
`lp_solver = "cpp_transport"`. The optional `"lp_transport"` and
`"lp_matrix"` backends require the suggested package `lpSolve`.

Partial transport does not infer the scientifically correct overlap for you:
choose `m` from the meaning of the data, or treat it as a sensitivity parameter
and report how conclusions change.

## When is the target distribution unknown?

Semirelaxed GW fixes the source marginal and learns the target marginal. This
fits problems where every source object must be represented but the appropriate
mass allocation over target objects is unknown.

```{r semirelaxed-fit}
semirelaxed <- entropic_semirelaxed_gromov_wasserstein(
  C1, C2, p = p, epsilon = 0.08, tol = 2e-8
)
data.frame(
  status = rfugw_status(semirelaxed),
  objective = rfugw_value(semirelaxed),
  target_mass = sum(rfugw_plan(semirelaxed))
)
```

```{r check-semirelaxed, include = FALSE}
semirelaxed_plan <- rfugw_plan(semirelaxed)
learned_q <- colSums(semirelaxed_plan)
stopifnot(
  all(is.finite(semirelaxed_plan)), all(semirelaxed_plan >= 0),
  diff(range(semirelaxed_plan)) > 0, diff(range(learned_q)) > 0,
  max(abs(rowSums(semirelaxed_plan) - p)) < 1e-6,
  abs(sum(learned_q) - 1) < 1e-6,
  is.finite(rfugw_value(semirelaxed)),
  identical(rfugw_status(semirelaxed), "converged")
)
```

This is not the same as partial transport: all source mass is transported, but
the solver decides how it is distributed across target columns.

## When may both marginals move?

`fugw_kl()` relaxes both marginal constraints using KL penalties. It is useful
when reference weights are meaningful but exact equality is too strong, for
example when collections contain unequal or unreliable mass.

```{r unbalanced-fit}
unbalanced <- fugw_kl(
  C1, C2, wx = p, wy = q, M = M,
  alpha = 0.5, epsilon = 0.5,
  reg_marginals = c(1, 1),
  max_iter = 200L, tol = 1e-6,
  max_iter_ot = 1000L, tol_ot = 1e-6
)
unbalanced
```

```{r check-unbalanced, include = FALSE}
unbalanced_plan <- rfugw_plan(unbalanced)
unbalanced_residuals <- rfugw_residuals(unbalanced)
stopifnot(
  all(is.finite(unbalanced_plan)), all(unbalanced_plan >= 0),
  diff(range(unbalanced_plan)) > 0, sum(unbalanced_plan) > 0,
  is.finite(rfugw_value(unbalanced)),
  identical(rfugw_status(unbalanced), "converged"),
  isTRUE(unbalanced$converged),
  identical(unbalanced_residuals$inner_status, "converged"),
  isTRUE(unbalanced_residuals$inner_converged)
)
```

Larger `reg_marginals` penalizes departure from `wx` and `wy` more strongly;
smaller values permit more creation, destruction, or redistribution of mass.
There is no universal setting: inspect the transported mass and marginal
deviations, and assess sensitivity on the scale relevant to your application.
The relatively smooth `epsilon` above keeps this small teaching example's
nested solve well conditioned; it is not a recommended value for every data
set.

## How do regularized and exact solvers differ?

Entropic solvers require a positive `epsilon` (or `reg` for the partial API).
Regularization smooths the coupling and changes the optimized objective. A
smaller value is not automatically better: it may make the inner transport
problem harder and can expose numerical limitations.

Exact conditional-gradient variants optimize the unregularized objective and
often yield sparser directions. They still solve a non-convex outer problem,
so neither family certifies a global optimum. For asymmetric structure costs,
leave `symmetric = NULL` so the solver detects the general path; setting
`symmetric = TRUE` requires genuinely symmetric inputs.

Start with documented defaults, normalize commensurate costs, and change one
control at a time. Use `sinkhorn_method = "log"` or stricter precision only
when diagnostics or the cost scale justify it; those are numerical choices,
not different scientific models.

## What should you check before using a plan?

For result objects, use the common interface:

```{r inspect-result}
rfugw_status(balanced)
rfugw_value(balanced)
rfugw_residuals(balanced)[c("feasible", "objective_consistent")]
```

Check four things:

1. The plan is finite, nonnegative, and non-degenerate.
2. Its marginals satisfy the formulation you selected.
3. Status and residuals support convergence rather than merely exhaustion of
   the iteration budget.
4. The objective is interpreted only within its exact formulation and
   regularization convention.

`ot_validate_plan()` checks balanced and partial marginal contracts. Unbalanced
plans have no fixed marginal target, so inspect their total mass, row and column
sums, and the diagnostics returned by `rfugw_residuals()`.

## Which APIs are experimental?

`sampled_gromov_wasserstein()`, `sampled_gromov_wasserstein_coords()`,
`sampled_gw_from_graphs()`, and `lowrank_gromov_wasserstein_samples()` are
experimental. Their names and signatures may change, and their result status
does not claim convergence certification. In particular, the low-rank samples
function factorizes a dense plan after fitting; it is not an end-to-end
low-memory GW solver.

Use these only when their documented envelope matches your task. See
`inst/bench/sampled-budget-curves.md` for the package's current evidence and
`inst/solver-contract.md` for detailed solver contracts.

## Where should you go next?

- `vignette("rfugw")` gives the shortest end-to-end FGW workflow.
- `vignette("linear-ot")` covers direct-cost balanced and unbalanced OT.
- `vignette("barycenters")` covers representative metric spaces.
- `vignette("multiset-alignment")` covers subject-to-template alignment.
