---
title: "Choosing the right GW solver"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Choosing the right GW solver}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 5,
  fig.height = 4
)
```

`rfugw` ships many Gromov-Wasserstein variants. This guide helps you pick the
right one for your problem.

## Setup: two small metric spaces

We'll use the same toy data throughout so you can compare outputs directly.

```{r data-setup}
library(rfugw)

set.seed(1)
ns <- 15
nt <- 18

# Source and target point clouds
X_src <- matrix(rnorm(ns * 2), ns, 2)
X_tgt <- matrix(rnorm(nt * 2), nt, 2)

# Structure matrices (normalized pairwise distances)
C1 <- as.matrix(dist(X_src)); C1 <- C1 / max(C1)
C2 <- as.matrix(dist(X_tgt)); C2 <- C2 / max(C2)

# Feature cost matrix for FGW variants
F_src <- matrix(rnorm(ns * 3), ns, 3)
F_tgt <- matrix(rnorm(nt * 3), nt, 3)
M <- as.matrix(dist(rbind(F_src, F_tgt)))[1:ns, (ns + 1):(ns + nt)]
M <- M / max(M)
```

## Decision tree

```
Do you have cross-domain features (M)?
├─ No  → GW (structure only)
│   ├─ Need exact solution?       → gromov_wasserstein()
│   └─ Speed matters?             → entropic_gromov_wasserstein()
└─ Yes → FGW (structure + features)
    ├─ Need exact solution?       → fgw_exact_cg()
    └─ Speed matters?             → fgw_entropic()

Are your distributions the same total mass?
├─ Yes, and both marginals fixed  → standard GW / FGW (above)
├─ Yes, but target mass is free   → semirelaxed variants
├─ No, different total mass       → partial or unbalanced variants
└─ No, and you want KL relaxation → fugw_kl()
```

## Standard GW and FGW

Both marginals are fixed (source and target weights sum to 1 each). The coupling
must satisfy `rowSums(plan) == p` and `colSums(plan) == q`.

```{r standard-gw}
# Exact GW (conditional gradient + LP direction)
out_gw <- gromov_wasserstein(C1, C2)
out_gw$gw_dist
```

```{r standard-fgw}
# Entropic FGW (the workhorse)
out_fgw <- fgw_entropic(M, C1, C2, alpha = 0.5, epsilon = 0.05)
out_fgw$fgw_dist
```

```{r exact-fgw}
# Exact FGW (conditional gradient, no regularization)
out_exact <- fgw_exact_cg(M, C1, C2, alpha = 0.5)
out_exact$fgw_dist
```

**When to use exact vs. entropic:** Exact solvers give sparser, more
interpretable couplings but scale poorly (LP subproblem at each iteration).
Entropic solvers are much faster for problems beyond ~50 points.

## Partial GW and FGW

Sometimes only a fraction of the mass should be transported---useful when the
two domains have different sizes or when you expect only partial overlap.

The `m` parameter controls how much mass to transport.

```{r partial-gw}
out_pgw <- partial_gromov_wasserstein(C1, C2, m = 0.5, log = TRUE)
out_pgw$partial_gw_dist

# Total transported mass
sum(out_pgw$plan)
```

```{r partial-fgw}
out_pfgw <- partial_fused_gromov_wasserstein(M, C1, C2, alpha = 0.5, m = 0.7, log = TRUE)
out_pfgw$partial_fgw_dist
```

Entropic partial variants are also available:

```{r entropic-partial}
out_epgw <- entropic_partial_gromov_wasserstein(C1, C2, m = 0.6, reg = 0.05, log = TRUE)
out_epgw$partial_gw_dist
```

## Semirelaxed GW and FGW

In semirelaxed mode, the source marginal is fixed but the target marginal is
*free*---the solver learns it. This is useful when you don't know the right
target distribution a priori (e.g., aligning a subject to a template of unknown
density).

```{r semirelaxed-gw}
out_sr <- semirelaxed_gromov_wasserstein(C1, C2)
out_sr$srgw_dist

# Source marginal is fixed
summary(rowSums(out_sr$plan))

# Target marginal was learned
round(out_sr$q, 4)
```

Entropic semirelaxed variants use mirror descent in KL geometry and are
significantly faster:

```{r entropic-semirelaxed}
out_esr <- entropic_semirelaxed_gromov_wasserstein(
  C1, C2,
  epsilon = 0.05
)
out_esr$srgw_dist
```

```{r entropic-semirelaxed-fgw}
out_esrf <- entropic_semirelaxed_fused_gromov_wasserstein(
  M, C1, C2,
  alpha = 0.5,
  epsilon = 0.05
)
out_esrf$srfgw_dist
```

## Fused Unbalanced GW (FUGW)

When both marginals can deviate from the prescribed distributions, use the
unbalanced solver. It relaxes the marginal constraints with KL divergence
penalties controlled by `reg_marginals`.

```{r fugw}
out_fugw <- fugw_kl(
  Cx = C1,
  Cy = C2,
  M = M,
  alpha = 0.5,
  epsilon = 0.01,
  reg_marginals = c(10, 10)
)
out_fugw$fugw_cost
```

Larger `reg_marginals` enforces the marginals more strictly (closer to standard
FGW). Smaller values allow more deviation.

```{r fugw-marginals}
# Compare marginals to uniform
cat("Source mass:", sum(out_fugw$pi_samp), "\n")
cat("Row marginal range:", range(rowSums(out_fugw$pi_samp)), "\n")
```

## Performance tips

### Mixed precision

For large problems, `precision = "mixed"` uses single-precision arithmetic for
inner Sinkhorn iterations and double for the final objective. This is the
default in `multialign_fit()` and typically 1.5--2x faster with negligible
accuracy loss.

```{r precision-compare}
t_double <- system.time(
  fgw_entropic(M, C1, C2, alpha = 0.5, epsilon = 0.05, precision = "double")
)
t_mixed <- system.time(
  fgw_entropic(M, C1, C2, alpha = 0.5, epsilon = 0.05, precision = "mixed")
)
cat("double:", t_double["elapsed"], "s, mixed:", t_mixed["elapsed"], "s\n")
```

### Check frequency

Increase `check_every` to reduce the frequency of convergence checks. Each
check involves computing a Frobenius norm, so larger values skip more work:

```{r check-every}
# Check convergence every 10 outer iterations instead of every 1
out <- fgw_entropic(
  M, C1, C2,
  alpha = 0.5,
  epsilon = 0.05,
  check_every = 10L
)
```

### Low-rank structure approximation

For very large structure matrices, the `structure_rank` parameter enables a
low-rank approximation of the GW tensor product, reducing cost from
O(n^2 * m) to O(n * m * r):

```{r low-rank}
out_lr <- fgw_entropic(
  M, C1, C2,
  alpha = 0.5,
  epsilon = 0.05,
  structure_rank = 5L
)
out_lr$fgw_dist
```

## Summary table

| Variant | Function | Marginals | Best for |
|---------|----------|-----------|----------|
| Standard GW | `gromov_wasserstein()` | Both fixed | Small exact problems |
| Entropic GW | `entropic_gromov_wasserstein()` | Both fixed | Fast structure matching |
| Standard FGW | `fgw_exact_cg()` | Both fixed | Exact structure + features |
| Entropic FGW | `fgw_entropic()` | Both fixed | Fast structure + features |
| Partial GW/FGW | `partial_*()` | Partial mass | Incomplete overlap |
| Semirelaxed GW/FGW | `semirelaxed_*()` | Source fixed, target free | Unknown target density |
| Unbalanced FUGW | `fugw_kl()` | Both relaxed (KL) | Noisy, unbalanced data |

## Next steps

- `vignette("barycenters")`: computing average shapes
- `vignette("multiset-alignment")`: aligning many subjects to a shared template
