---
title: A solver-client protocol for downstream libraries
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: structural
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{A solver-client protocol for downstream libraries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE, warning = FALSE)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

The solver-client protocol gives an R package one stable route from an
explicit transport estimand to a certified coupling. The generic example
comes first here: DKGE and manifoldalign use the same boundary, but neither
package defines its semantics.

```{r load}
library(rfugw)
```

## What data and estimand are we declaring?

Suppose a client has seven source objects, five target objects, a rectangular
cross-cost matrix, and probability weights. Rows of the cost are always source
support; columns are target support. Our estimand is balanced entropic linear
transport, and the expected output is a 7-by-5 coupling with the supplied row
and column marginals.

```{r problem}
set.seed(31)
source_points <- matrix(rnorm(21), 7, 3)
target_points <- matrix(rnorm(15), 5, 3)
cost <- outer(rowSums(source_points^2), rowSums(target_points^2), "+") -
  2 * tcrossprod(source_points, target_points)
source_mass <- seq_len(7) / sum(seq_len(7))
target_mass <- rev(seq_len(5)) / sum(seq_len(5))

problem <- transport_problem_sinkhorn(
  cost,
  p = source_mass,
  q = target_mass,
  epsilon = 0.7,
  method = "auto",
  tol = 1e-9,
  mass_policy = "probability"
)
problem
```

`mass_policy = "probability"` verifies rather than silently rescales. Use
`"normalize"` only when separate normalization is the intended estimand; the
problem then retains both original and effective masses. KL-unbalanced
problems instead require an explicit finite-measure, joint, or separate policy.

## How does a client solve and inspect it?

```{r solve}
fit <- transport_solve(problem)
plan <- rfugw_plan(fit)

data.frame(
  status = rfugw_status(fit),
  rows = nrow(plan),
  columns = ncol(plan),
  row_residual = rfugw_residuals(fit)$row_residual,
  column_residual = rfugw_residuals(fit)$col_residual
)
```

The public accessors separate four concerns:

- `rfugw_problem()` returns the declared estimand and requested controls;
- `rfugw_masses()` separates input, effective, and transported mass;
- `rfugw_provenance()` records requested/effective backend controls, objective
  units, representation, status, and runtime;
- the established plan, value, status, and residual accessors remain the
  authoritative numerical boundary.

```{r provenance}
masses <- rfugw_masses(fit)
provenance <- rfugw_provenance(fit)
list(
  input_mass = c(
    source = masses$source$input_mass,
    target = masses$target$input_mass
  ),
  transported_mass = masses$transported_mass,
  estimand = provenance$estimand,
  backend = provenance$backend,
  objective_units = provenance$objective$units
)
```

## How is solver state reused safely?

`rfugw_state()` returns opaque versioned state only from a certified result.
Balanced state contains canonical dual potentials, so the next solve can
change tolerance, epsilon, or switch between scaling and log backends. It is
an initialization—not a cached result—and the continuation must earn a new
certificate.

```{r continuation}
state <- rfugw_state(fit)
tighter_problem <- transport_problem_sinkhorn(
  cost,
  p = source_mass,
  q = target_mass,
  epsilon = 0.55,
  method = "log",
  tol = 1e-10,
  mass_policy = "probability"
)
continued <- transport_solve(tighter_problem, init_state = state)

data.frame(
  status = rfugw_status(continued),
  warm_started = continued$warm_started,
  state_accepted = continued$warm_start_accepted,
  iterations = continued$iterations,
  residual = continued$residual
)
```

Invalid dimensions, nonfinite potentials, changed support, incompatible state
types, and unsafe scaling starts fail before computation. A client that wants
an explicit fallback can request `state_policy = "cold"`; the result then
contains a structured rejection code and does not claim it was warm-started.
Partial-Sinkhorn state is stricter than balanced state because its Dykstra
corrections are bound to the identical problem and effective backend.

State survives `saveRDS()` / `readRDS()`. The retained fresh-process installed
check is `inst/bench/verify_solver_protocol_state_roundtrip.R`.

## Does the same boundary preserve a second formulation?

Exact transport uses a different constructor and objective certificate. It
does not accept iterative state.

```{r exact}
exact_problem <- transport_problem_emd(
  cost,
  p = source_mass,
  q = target_mass,
  mass_policy = "probability"
)
exact <- transport_solve(exact_problem)
data.frame(
  estimand = rfugw_provenance(exact)$estimand,
  status = rfugw_status(exact),
  value = rfugw_value(exact),
  primal_dual_gap = exact$duality_gap
)
```

The formulation-specific constructor is important: exact OT, partial OT,
KL-unbalanced OT, GW, FGW, and FUGW do not become equivalent because their
results share accessors.

## What does protocol v1 support?

```{r capabilities}
transport_capabilities()[, c(
  "estimand", "protocol_constructor", "mass_policy", "warm_state",
  "plan_representation", "relational_costs"
)]
```

A false `protocol_constructor` means the distinct named public solver remains
available but is not wrapped by protocol v1. Sparse and implicit couplings are
consumed through the transport-plan/operator API; materialization stays an
explicit action. Application-specific intensive versus extensive
normalization remains in the downstream library.

The installed generic fixture is
`inst/extdata/clients/generic_solver_protocol_fixture.R`. A DKGE-shaped
fixture and differential test reproduce its certified log-Sinkhorn plan,
objective, marginal error, failure behavior, and tighter-tolerance reuse using
only exported rfugw APIs on the client side. Manifoldalign fixtures exercise
the same principle for OT-Procrustes and sparse translation-invariant UOT.
